---
title: "Final Project - Predicting PDAC with Circulating miRNA - Multi-miRNA and Consensus Feature Analysis"
authors: "Madison Blake, Kristyn Gumpper-Fedus, Sean Kirkpatrick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	dev = "png",
	fig.path = "Figures/",
	fig.retina = 1
)

dir.create("Data", showWarnings = FALSE, recursive = TRUE)
dir.create("Tables", showWarnings = FALSE, recursive = TRUE)
dir.create("Figures", showWarnings = FALSE, recursive = TRUE)

library(dplyr)
library(ggplot2)
library(readr)

source("R/plotHelpers.R")
source("R/multi_mir_analysis.R")

report_start_time <- Sys.time()

write_comma_separated_list <- function(items) {
	if (length(items) == 0) {
		return("")
	} else if (length(items) == 1) {
		return(items)
	} else {
		return(paste(
			paste(items[-length(items)], collapse = ", "),
			items[length(items)],
			sep = ifelse(length(items) > 2, ", and ", " and ")
		))
	}
}

first_non_missing <- function(values) {
	values <- values[!is.na(values) & values != ""]

	if (length(values)) {
		values[[1]]
	} else {
		NA_character_
	}
}

```

```{r report_table_styles, echo=FALSE, results='asis'}

cat(
	paste(
		"<style>",
		"table {",
		"  border-collapse: collapse;",
		"}",
		"table th,",
		"table td {",
		"  border: 1px solid #cfcfcf;",
		"  padding: 4px 8px;",
		"}",
		"</style>",
		sep = "\n"
	)
)

```

```{r constants, include=FALSE}

gene_family_file <- "Data/GSE314577_family.xml/GPL30934-tbl-1.txt"
downstream_analysis_inputs_rds_file <- "Data/Top-ranked consensus miRNA downstream inputs.rds"
run_multimir_analysis <- TRUE
multimir_results_rds_file <- "Data/Top-ranked consensus miRNA multiMiR results.rds"

```

```{r load_downstream_inputs, include=FALSE}

if (!file.exists(downstream_analysis_inputs_rds_file)) {
	stop(
		paste(
			"Downstream analysis input file not found:",
			downstream_analysis_inputs_rds_file,
			"Run PDACmiRNAAnalysis.rmd first to generate the saved consensus-feature inputs."
		),
		call. = TRUE
	)
}

downstream_analysis_inputs <- readRDS(downstream_analysis_inputs_rds_file)
required_input_names <- c("top_ranked_miRNA_details", "merged_top_features")
missing_input_names <- setdiff(required_input_names, names(downstream_analysis_inputs))

if (length(missing_input_names) > 0) {
	stop(
		paste(
			"Downstream analysis input file is missing required objects:",
			paste(missing_input_names, collapse = ", ")
		),
		call. = TRUE
	)
}

top_ranked_miRNA_details <- downstream_analysis_inputs$top_ranked_miRNA_details
merged_top_features <- unique(as.character(downstream_analysis_inputs$merged_top_features))

if (!is.data.frame(top_ranked_miRNA_details) || nrow(top_ranked_miRNA_details) == 0) {
	stop("Saved top_ranked_miRNA_details is missing or empty.", call. = TRUE)
}

if (!length(merged_top_features)) {
	stop("Saved merged_top_features is missing or empty.", call. = TRUE)
}

gpl <- read_tsv(
	gene_family_file,
	col_names = c("Annotation", "Alias", "MIMAT", "ClassName"),
	show_col_types = FALSE
)

annotation_lookup <- gpl %>%
	distinct(.data$MIMAT, .keep_all = TRUE) %>%
	transmute(
		miRNA = .data$MIMAT,
		Annotation = .data$Annotation
	)

merged_top_feature_details <- data.frame(
	miRNA = merged_top_features,
	n_models_selected = 1L,
	stringsAsFactors = FALSE
) %>%
	left_join(annotation_lookup, by = "miRNA")

merged_top_feature_details$multiMiR_query_id <- vapply(
	merged_top_feature_details$Annotation,
	FUN.VALUE = character(1),
	function(annotation_value) {
		if (is.na(annotation_value) || !nzchar(annotation_value)) {
			return(NA_character_)
		}

		trimws(strsplit(annotation_value, ",", fixed = TRUE)[[1]][[1]])
	}
)

write.csv(
	top_ranked_miRNA_details,
	"Tables/Top-ranked Consensus miRNA Details.csv",
	row.names = FALSE
)

write.csv(
	merged_top_feature_details,
	"Tables/Merged Top miRNA Features for Downstream multiMiR Queries.csv",
	row.names = FALSE
)

top_rank_count <- suppressWarnings(max(top_ranked_miRNA_details$n_models_selected, na.rm = TRUE))

if (!is.finite(top_rank_count)) {
	top_rank_count <- 0L
}

```

# Multi-miRNA and Consensus Feature Analysis

This report loads the saved consensus-feature inputs written by the main PDAC modeling report and performs the downstream multiMiR and final consensus-feature summary analysis separately.

## Top-ranked Consensus miRNAs

```{r top_ranked_consensus_mirna, results='asis'}

cat(
	"<p>Loaded downstream consensus-feature inputs from ",
	downstream_analysis_inputs_rds_file,
	". Top-rank selection count: ",
	top_rank_count,
	" non-tree models.</p>\n",
	sep = ""
)

render_html_table(
	top_ranked_miRNA_details %>%
		dplyr::select(
			miRNA,
			Annotation,
			n_models_selected,
			limma_log_fc,
			limma_t,
			limma_p_value,
			limma_adj_p_value
		),
	caption = "Top-ranked consensus miRNAs from the non-tree models"
)

if (nrow(top_ranked_miRNA_details) > 1) {
	cat(
		paste0(
			"<p>Multiple miRNAs tied for the highest non-tree model-selection count, ",
			"so the table is ordered by the smallest blocked-limma p-value, ",
			"with ties broken by the absolute moderated t-statistic and then miRNA identifier. ",
			"All rows in this table feed the downstream consensus-feature summary outputs.</p>\n"
		)
	)
}

```

## Merged miRNA Set for Downstream multiMiR Queries

```{r merged_top_feature_inputs, results='asis'}

cat(
	"<p>The downstream multiMiR query set contains ",
	nrow(merged_top_feature_details),
	" merged top miRNA features drawn from the consensus, decision-tree, and random-forest rankings.</p>\n",
	sep = ""
)

render_html_table(
	merged_top_feature_details %>%
		dplyr::select(
			miRNA,
			Annotation,
			multiMiR_query_id
		),
	caption = "Merged top miRNA features with derived multiMiR query identifiers"
)

```

## miRNA Interactions

Assess validated target and disease/drug associations for the merged top-ranked miRNA set with multiMiR.

```{r multi_mir_analysis, results='asis'}

if (run_multimir_analysis) {
	multimir_results <- run_multi_mir_queries(merged_top_feature_details)
	saveRDS(multimir_results, multimir_results_rds_file)
} else {
	if (!file.exists(multimir_results_rds_file)) {
		stop(
			paste(
				"Saved multiMiR results file not found:",
				multimir_results_rds_file,
				"Set run_multimir_analysis <- TRUE to refresh the downstream multiMiR results."
			),
			call. = TRUE
		)
	}

	multimir_results <- readRDS(multimir_results_rds_file)

	cat(
		"<p>Loaded saved multiMiR results from ",
		multimir_results_rds_file,
		" instead of issuing live multiMiR queries.</p>\n",
		sep = ""
	)
}

if (nrow(multimir_results$query_failures) > 0) {
	cat(
		"<p>",
		nrow(multimir_results$query_failures),
		" multiMiR query call",
		if (nrow(multimir_results$query_failures) == 1) " was" else "s were",
		" skipped after returning an error. The report will continue with any successful multiMiR results.</p>\n",
		sep = ""
	)
}

if (nrow(multimir_results$queried_miRNA) == 0) {
	cat("<p>No valid miRNA annotation was available for the merged top-feature set, so multiMiR queries were skipped.</p>\n")
} else {
	cat(
		"<p>multiMiR query miRNAs: ",
		write_comma_separated_list(multimir_results$queried_miRNA$multiMiR_query_id),
		"</p>\n",
		sep = ""
	)

	if (multimir_results$skipped_query_count > 0) {
		cat(
			"<p>",
			multimir_results$skipped_query_count,
			" merged top-feature miRNA",
			if (multimir_results$skipped_query_count == 1) " was" else "s were",
			" skipped because no valid multiMiR query identifier was available.</p>\n",
			sep = ""
		)
	}

	validated_target_summary <- multimir_results$validated$summary

	if (!is.null(validated_target_summary) && nrow(validated_target_summary) > 0) {
		write.csv(
			validated_target_summary,
			"Tables/Top-ranked Consensus miRNA multiMiR Validated Targets.csv",
			row.names = FALSE
		)

		unique_validated_targets <- unique(
			validated_target_summary$target_symbol[
				!is.na(validated_target_summary$target_symbol) &
					validated_target_summary$target_symbol != ""
			]
		)

		cat(
			"<p>Validated multiMiR interaction records across the merged top-feature miRNA set: ",
			multimir_results$validated$total_record_count,
			"; unique target genes: ",
			length(unique_validated_targets),
			"; queried miRNAs with validated targets: ",
			length(unique(validated_target_summary$miRNA)),
			" of ",
			nrow(multimir_results$queried_miRNA),
			"</p>\n",
			sep = ""
		)

		render_html_table(
			utils::head(validated_target_summary, 15),
			caption = "Top validated multiMiR targets for the merged top-feature miRNA set"
		)
	} else {
		cat("<p>No validated multiMiR target interactions were returned for the merged top-feature miRNA set.</p>\n")
	}

	disease_association_summary <- multimir_results$disease_drug$summary

	if (!is.null(disease_association_summary) && nrow(disease_association_summary) > 0) {
		write.csv(
			disease_association_summary,
			"Tables/Top-ranked Consensus miRNA multiMiR Disease-Drug Associations.csv",
			row.names = FALSE
		)

		unique_association_terms <- unique(
			disease_association_summary$disease_drug[
				!is.na(disease_association_summary$disease_drug) &
					disease_association_summary$disease_drug != ""
			]
		)

		cat(
			"<p>multiMiR disease/drug association records across the merged top-feature miRNA set: ",
			multimir_results$disease_drug$total_record_count,
			"; unique association terms: ",
			length(unique_association_terms),
			"; queried miRNAs with disease/drug associations: ",
			length(unique(disease_association_summary$miRNA)),
			" of ",
			nrow(multimir_results$queried_miRNA),
			"</p>\n",
			sep = ""
		)

		render_html_table(
			utils::head(disease_association_summary, 15),
			caption = "Top multiMiR disease/drug associations for the merged top-feature miRNA set"
		)
	} else {
		cat("<p>No multiMiR disease/drug associations were returned for the merged top-feature miRNA set.</p>\n")
	}
}

```

## Shared Validated multiMiR Target Genes

Summarize validated target genes across the saved multiMiR results object for the merged top-feature miRNA set.

```{r multimir_gene_summary, results='asis'}

multimir_gene_summary <- NULL

validated_target_summary <- multimir_results$validated$summary

if (is.null(validated_target_summary) || nrow(validated_target_summary) == 0) {
	cat(
		"<p>The saved multiMiR results did not contain any validated target summary rows.</p>\n"
	)
} else {
	multimir_gene_summary <- validated_target_summary %>%
		mutate(
			miRNA_label = if_else(
				!is.na(.data$Annotation) & .data$Annotation != "",
				.data$Annotation,
				.data$miRNA
			)
		) %>%
		group_by(.data$target_symbol) %>%
		summarise(
			target_entrez = first_non_missing(.data$target_entrez),
			target_ensembl = first_non_missing(.data$target_ensembl),
			n_consensus_miRNA = n_distinct(.data$miRNA),
			consensus_miRNA = write_comma_separated_list(sort(unique(.data$miRNA_label))),
			total_validated_support = sum(.data$validated_support, na.rm = TRUE),
			total_support = sum(.data$total_support, na.rm = TRUE),
			max_models_selected = max(.data$n_models_selected, na.rm = TRUE),
			.groups = "drop"
		) %>%
		arrange(
			desc(.data$n_consensus_miRNA),
			desc(.data$total_validated_support),
			desc(.data$total_support),
			.data$target_symbol
		)

	write.csv(
		multimir_gene_summary,
		"Tables/Top-ranked Consensus miRNA multiMiR Target Gene Summary.csv",
		row.names = FALSE
	)

	cat(
		"<p>Loaded saved multiMiR results from ",
		multimir_results_rds_file,
		" and summarized ",
		nrow(multimir_gene_summary),
		" validated target genes across ",
		nrow(multimir_results$queried_miRNA),
		" queried merged top-feature miRNAs.</p>\n",
		sep = ""
	)

	render_html_table(
		utils::head(multimir_gene_summary, 20),
		caption = "Validated target genes shared across the merged top-feature miRNA set"
	)
}

```

## Shared multiMiR Disease/Drug Terms

Summarize disease/drug terms across the saved multiMiR results object for the merged top-feature miRNA set.

```{r multimir_disease_drug_top_terms, results='asis'}

disease_association_summary <- multimir_results$disease_drug$summary

if (is.null(disease_association_summary) || nrow(disease_association_summary) == 0) {
	cat(
		"<p>The saved multiMiR results did not contain any disease/drug association summary rows.</p>\n"
	)
} else {
	multimir_disease_drug_summary <- disease_association_summary %>%
		mutate(
			miRNA_label = if_else(
				!is.na(.data$Annotation) & .data$Annotation != "",
				.data$Annotation,
				.data$miRNA
			),
			disease_drug = trimws(.data$disease_drug)
		) %>%
		filter(!is.na(.data$disease_drug), .data$disease_drug != "") %>%
		group_by(.data$disease_drug) %>%
		summarise(
			n_consensus_miRNA = n_distinct(.data$miRNA),
			consensus_miRNA = write_comma_separated_list(sort(unique(.data$miRNA_label))),
			total_records = sum(.data$n_records, na.rm = TRUE),
			total_databases = sum(.data$n_databases, na.rm = TRUE),
			total_pubmed = sum(.data$n_pubmed, na.rm = TRUE),
			max_models_selected = max(.data$n_models_selected, na.rm = TRUE),
			.groups = "drop"
		) %>%
		arrange(
			desc(.data$n_consensus_miRNA),
			desc(.data$total_records),
			desc(.data$total_pubmed),
			.data$disease_drug
		)

	write.csv(
		multimir_disease_drug_summary,
		"Tables/Top-ranked Consensus miRNA multiMiR Disease-Drug Term Summary.csv",
		row.names = FALSE
	)

	cat(
		"<p>Loaded saved multiMiR results from ",
		multimir_results_rds_file,
		" and summarized ",
		nrow(multimir_disease_drug_summary),
		" disease/drug terms across ",
		nrow(multimir_results$queried_miRNA),
		" queried merged top-feature miRNAs.</p>\n",
		sep = ""
	)

	render_html_table(
		multimir_disease_drug_summary,
		caption = "Disease/drug terms shared across the merged top-feature miRNA set"
	)
}

```

# Final Summary Table of the Top-Ranked Consensus miRNAs

A table of the top-ranked consensus miRNAs with their blocked-limma statistics, model selection counts, and multiMiR summary counts.

```{r summary_table_consensus_features, results='asis'}

validated_target_counts <- data.frame(
	miRNA = character(0),
	n_target_genes = integer(0),
	stringsAsFactors = FALSE
)

if (!is.null(multimir_results$validated$summary) && nrow(multimir_results$validated$summary) > 0) {
	validated_target_counts <- multimir_results$validated$summary %>%
		filter(!is.na(.data$target_symbol) & .data$target_symbol != "") %>%
		group_by(.data$miRNA) %>%
		summarise(
			n_target_genes = n_distinct(.data$target_symbol),
			.groups = "drop"
		)
}

disease_drug_counts <- data.frame(
	miRNA = character(0),
	n_disease_drug_terms = integer(0),
	stringsAsFactors = FALSE
)

if (!is.null(multimir_results$disease_drug$summary) && nrow(multimir_results$disease_drug$summary) > 0) {
	disease_drug_counts <- multimir_results$disease_drug$summary %>%
		filter(!is.na(.data$disease_drug) & .data$disease_drug != "") %>%
		group_by(.data$miRNA) %>%
		summarise(
			n_disease_drug_terms = n_distinct(.data$disease_drug),
			.groups = "drop"
		)
}

final_feature_table <- top_ranked_miRNA_details %>%
	dplyr::select(
		miRNA,
		Annotation,
		mean_control,
		mean_case,
		mean_diff,
		limma_p_value,
		n_models_selected
	) %>%
	left_join(validated_target_counts, by = "miRNA") %>%
	left_join(disease_drug_counts, by = "miRNA") %>%
	mutate(
		n_target_genes = dplyr::coalesce(.data$n_target_genes, 0L),
		n_disease_drug_terms = dplyr::coalesce(.data$n_disease_drug_terms, 0L)
	)

write.csv(final_feature_table, "Tables/Summary of Consensus miRNAs.csv", row.names = FALSE)

render_html_table(
	final_feature_table,
	caption = "Top-ranked consensus miRNAs summary",
	digits = 3
)

```

## Blocked-limma Summary Plot of Top-Ranked Consensus miRNAs

Blocked-limma summary of the top-ranked consensus miRNAs.

```{r final_limma_summary_plot, results='asis', fig.width=7, fig.height=5}

limma_plot_data <- top_ranked_miRNA_details %>%
	mutate(
		plot_p_value = pmax(.data$limma_p_value, .Machine$double.xmin),
		plot_label = if_else(!is.na(.data$Annotation) & .data$Annotation != "", .data$Annotation, .data$miRNA)
	)

limma_summary_plot <- ggplot(
	limma_plot_data,
	aes(x = limma_log_fc, y = -log10(plot_p_value), color = limma_t)
) +
	geom_point(size = 3) +
	geom_text(aes(label = plot_label), nudge_y = 0.1, check_overlap = TRUE, size = 3) +
	scale_color_gradient2(low = "blue", mid = "grey90", high = "red", midpoint = 0, na.value = "grey") +
	labs(
		title = "Blocked-limma Summary of Top-Ranked Consensus miRNAs",
		x = "Limma logFC",
		y = "-Log10 limma P-Value",
		color = "Limma t"
	) +
	theme_minimal()

association_plot_file <- "Figures/Blocked Limma Summary Top Consensus miRNAs.png"

save_plot(
	filename = association_plot_file,
	plot = limma_summary_plot
)

cat(
	sprintf(
		'<img src="%s" alt="Blocked-limma Summary of Top-Ranked Consensus miRNAs" style="width:100%%; max-width:700px;" />\n\n',
		association_plot_file
	)
)

```

```{r final_report_conclusion, results='asis'}

report_end_time <- Sys.time()
elapsed_seconds <- as.numeric(difftime(report_end_time, report_start_time, units = "secs"))

format_elapsed_time <- function(elapsed_seconds) {
	total_seconds <- max(0L, as.integer(round(elapsed_seconds)))
	hours <- total_seconds %/% 3600L
	minutes <- (total_seconds %% 3600L) %/% 60L
	seconds <- total_seconds %% 60L

	sprintf("%02d:%02d:%02d", hours, minutes, seconds)
}

cat(
	paste0(
		"<p>Report runtime: ",
		format_elapsed_time(elapsed_seconds),
		" (started ",
		format(report_start_time, "%Y-%m-%d %H:%M:%S"),
		"; finished ",
		format(report_end_time, "%Y-%m-%d %H:%M:%S"),
		").</p>\n"
	)
)

```

