---
title: "Final Project - Predicting PDAC with Circulating miRNA"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

setwd("~/Classes/BMI7235 - Machine Learning/Final Project")

library(readxl)
library(readr)
library(tidyr)
library(glmnet)
library(ggplot2)
library(pROC)
library(dplyr)
library(rpart)
library(randomForest)

source("R/run_fold_lasso.R")
source("R/run_fold_enet.R")
source("R/run_fold_forward_step.R")
source("R/run_fold_lda.R")
source("R/run_fold_rf.R")
source("R/run_fold_dt.R")

render_html_table <- function(data, caption = NULL, digits = NULL) {
  kable_args <- list(
    x = data,
    format = "html",
    row.names = FALSE,
    caption = caption
  )

  if (!is.null(digits)) {
    kable_args$digits <- digits
  }

  cat(
    do.call(knitr::kable, kable_args),
    sep = "\n"
  )
}

model_display_name <- function(model_name) {
  tools::toTitleCase(gsub("_", " ", model_name))
}

```

```{r constants, include=FALSE}
# constants and model parameters

#random seed
set.seed(31415)

model_names <- c("forward_step", "lasso", "enet", "lda", "random_forest", "decision_tree")
enet_alpha_grid <- c(0.1, 0.3, 0.5, 0.7, 0.9)
forward_step_max_steps <- 20
lasso_alpha <- 1
lda_prior <- c(0.5, 0.5)
rf_ntree <- 1000
dt_maxdepth <- 4

fold_filter_name <- "class SNR"
fold_filter_top_n <- 50
interpretation_method_name <- "per-miRNA logistic regression on PDAC status"
filter_stat_cols <- c("miRNA", "mean_control", "mean_case", "mean_diff", "sd_control", "sd_case", "class_snr", "p_value")
association_cols <- c("miRNA", "estimate", "odds_ratio", "std_error", "z_value", "p_value", "adj_p_value")

```


# Preparing Data

## Import Data
We use the normalized miRNA count matrix from GEO and merge it with the demographic metadata.  Categorical variables are set to factors and the PDAC case and control variable is set to binary for modeling.

```{r, include=FALSE}

normalized_counts <- read_xlsx("GSE314577/GSE314577_VUMC_miRNA_normalized_data.xlsx")
metadata <- read_xlsx("GSE314577/GSE314577_VUMC_Metadata_ch_revised.xlsx")

normalized_counts <- normalized_counts %>%
  mutate(ID = as.character(.data$ID))

metadata <- metadata %>%
  mutate(
    ID = as.character(.data$ID),
    CC = factor(.data$CC, levels = c("Controls", "PDAC case")),
    Sex = factor(.data$Sex, levels = c("Female", "Male")),
    Race = factor(.data$Race, levels = c("Asian", "Black", "White", "Others")),
    matched_set = factor(.data$matched_set),
    Age = as.numeric(.data$age_sam),
    CC_bin = ifelse(.data$CC == "PDAC case", 1, 0)
  )

full_dat <- metadata %>%
  left_join(normalized_counts, by = "ID")

miRNA_cols <- grep("^MIMAT", names(full_dat), value = TRUE)

cat("Rows/columns in merged dataset:", dim(full_dat), "\n")
cat("Number of miRNA columns:", length(miRNA_cols), "\n")

```

## Create Training and Holdout Sets
To split the data while respecting matched-pairs, we performed the split on the matched set IDs, ensuring each pair is not split between sets.

```{r, echo=FALSE, results='asis'}

all_sets <- unique(full_dat$matched_set)
n_train_sets <- floor(0.8 * length(all_sets))
train_set_ids <- sample(all_sets, n_train_sets)

train_set <- full_dat %>% filter(matched_set %in% train_set_ids)
holdout  <- full_dat %>% filter(!matched_set %in% train_set_ids)

split_summary <- data.frame(
  dataset = c("Training", "Holdout"),
  rows = c(nrow(train_set), nrow(holdout)),
  row.names = NULL
)

class_balance <- bind_rows(
  data.frame(
    dataset = "Training",
    class = names(table(train_set$CC)),
    n_samples = as.integer(table(train_set$CC)),
    row.names = NULL
  ),
  data.frame(
    dataset = "Holdout",
    class = names(table(holdout$CC)),
    n_samples = as.integer(table(holdout$CC)),
    row.names = NULL
  )
)

render_html_table(split_summary, caption = "Training and holdout set sizes")
render_html_table(class_balance, caption = "Training and holdout class balance")

```

# Cross-Validation Loop
We perform a k-fold cross-validation on the training set of paired samples, ensuring matched-samples are kept together in the same fold.

Within each fold:
1. We apply a fold-specific feature filter, currently class-based signal-to-noise ratio (SNR) on the normalized counts, to select the top miRNAs within the training fold.
2. We then fit each model using only the selected miRNAs from the normalized count matrix.

The SNR filtering is calculated as the absolute difference in means between cases and controls divided by the sum of standard deviations, with an additional t-test for association. The top N miRNAs are selected based on the highest SNR values, with ties broken by p-value.
```{r, include=FALSE}

# Feature filter
#  The top N miRNAs are selected based on the highest SNR values, with ties broken by p-value from two-sample t-test
fold_filter_fn <- function(train_fold_dat, miRNA_cols, top_n = 50) {
  case_stats <- train_fold_dat %>%
    filter(.data$CC == "PDAC case") %>%
    summarise(
      across(
        tidyselect::all_of(miRNA_cols),
        list(
          mean = ~ mean(.x, na.rm = TRUE),
          sd = ~ sd(.x, na.rm = TRUE)
        )
      )
    ) %>%
    pivot_longer(
      cols = everything(),
      names_to = c("miRNA", ".value"),
      names_sep = "_"
    ) %>%
    rename(
      mean_case = mean,
      sd_case = sd
    )

  control_stats <- train_fold_dat %>%
    filter(.data$CC == "Controls") %>%
    summarise(
      across(
        tidyselect::all_of(miRNA_cols),
        list(
          mean = ~ mean(.x, na.rm = TRUE),
          sd = ~ sd(.x, na.rm = TRUE)
        )
      )
    ) %>%
    pivot_longer(
      cols = everything(),
      names_to = c("miRNA", ".value"),
      names_sep = "_"
    ) %>%
    rename(
      mean_control = mean,
      sd_control = sd
    )

  p_values <- bind_rows(lapply(miRNA_cols, function(miRNA_name) {
    data.frame(
      miRNA = miRNA_name,
      p_value = tryCatch(
        t.test(train_fold_dat[[miRNA_name]] ~ train_fold_dat$CC)$p.value,
        error = function(e) NA_real_
      ),
      stringsAsFactors = FALSE
    )
  }))

  class_stats <- case_stats %>%
    left_join(control_stats, by = "miRNA") %>%
    left_join(p_values, by = "miRNA") %>%
    mutate(
      mean_diff = .data$mean_case - .data$mean_control,
      class_snr = if_else(
        (.data$sd_case + .data$sd_control) > 0,
        abs(.data$mean_diff) / (.data$sd_case + .data$sd_control),
        NA_real_
      )
    ) %>%
    filter(is.finite(.data$class_snr), is.finite(.data$p_value)) %>%
    select(
      tidyselect::all_of(c(
        "miRNA",
        "mean_control",
        "mean_case",
        "mean_diff",
        "sd_control",
        "sd_case",
        "class_snr",
        "p_value"
      ))
    ) %>%
    arrange(desc(.data$class_snr), .data$p_value)

  selected <- class_stats %>%
    slice_head(n = top_n) %>%
    pull("miRNA")

  list(
    stats = class_stats,
    selected = selected
  )
}

```

```{r, include=FALSE}
prepare_fold_inputs <- function(train_fold_dat, valid_fold_dat, selected_miRNA) {
  train_fold_dat <- train_fold_dat %>%
    mutate(CC_bin = if_else(.data$CC == "PDAC case", 1, 0))
  valid_fold_dat <- valid_fold_dat %>%
    mutate(CC_bin = if_else(.data$CC == "PDAC case", 1, 0))

  if (!length(selected_miRNA)) {
    stop("Feature filtering returned zero miRNAs.", call. = TRUE)
  }

  complete_feature_flags <- bind_rows(
    train_fold_dat %>%
      summarise(
        across(
          tidyselect::all_of(selected_miRNA),
          ~ all(!is.na(.x))
        )
      ) %>%
      mutate(dataset = "train"),
    valid_fold_dat %>%
      summarise(
        across(
          tidyselect::all_of(selected_miRNA),
          ~ all(!is.na(.x))
        )
      ) %>%
      mutate(dataset = "valid")
  )

  selected_miRNA <- complete_feature_flags %>%
    pivot_longer(
      cols = tidyselect::all_of(selected_miRNA),
      names_to = "miRNA",
      values_to = "is_complete"
    ) %>%
    group_by(.data$miRNA) %>%
    summarise(is_complete = all(.data$is_complete), .groups = "drop") %>%
    filter(.data$is_complete) %>%
    pull("miRNA")

  if (!length(selected_miRNA)) {
    stop("All selected miRNAs were removed after screening for complete data in both training and evaluation sets.", call. = TRUE)
  }

#TODO:  add demographic variables to the model and the variable 

  keep_cols <- c("CC", "CC_bin", selected_miRNA)
  train_model_dat <- train_fold_dat %>%
    select(tidyselect::all_of(keep_cols))
  valid_model_dat <- valid_fold_dat %>%
    select(tidyselect::all_of(keep_cols))
  model_formula <- stats::reformulate(selected_miRNA, response = "CC_bin")
  class_formula <- stats::reformulate(selected_miRNA, response = "CC")

  list(
    selected_miRNA = selected_miRNA,
    model_formula = model_formula,
    class_formula = class_formula,
    x_train = stats::model.matrix(model_formula, data = train_model_dat)[, -1, drop = FALSE],
    x_valid = stats::model.matrix(model_formula, data = valid_model_dat)[, -1, drop = FALSE],
    y_train = train_model_dat$CC_bin,
    train_model_dat = train_model_dat,
    valid_model_dat = valid_model_dat
  )
}
```

```{r, include=FALSE}

cv_set_ids <- unique(train_set$matched_set)
k <- 5
fold_id <- sample(rep(1:k, length.out = length(cv_set_ids)))

fold_map <- data.frame(
  matched_set = cv_set_ids,
  fold = fold_id
)

cv_predictions <- list()
feature_tracker <- list()
tuning_tracker <- list()
feature_count_tracker <- list()
filter_sizes <- integer(k)

for (model_name in model_names) {
  cv_predictions[[model_name]] <- list()
  feature_tracker[[model_name]] <- list()
  tuning_tracker[[model_name]] <- list()
  feature_count_tracker[[model_name]] <- integer(k)
}

for (fold in 1:k) {
  cat("Running fold", fold, "of", k, "...\n")

  test_sets <- fold_map$matched_set[fold_map$fold == fold]

  cv_train <- train_set %>% filter(!matched_set %in% test_sets)
  cv_test <- train_set %>% filter(matched_set %in% test_sets)

  filter_out <- fold_filter_fn(
    train_fold_dat = cv_train,
    miRNA_cols = miRNA_cols,
    top_n = fold_filter_top_n
  )
  fold_inputs <- prepare_fold_inputs(
    cv_train,
    cv_test,
    filter_out$selected
  )

  filter_sizes[fold] <- length(filter_out$selected)

  model_outputs <- list(
    forward_step = run_fold_forward_step(
      fold_inputs,
      max_steps = forward_step_max_steps
    ),
    lasso = run_fold_lasso(
      fold_inputs,
      alpha_lasso = lasso_alpha
    ),
    enet = run_fold_enet(
      fold_inputs,
      alpha_enet = enet_alpha_grid
    ),
    lda = run_fold_lda(
      fold_inputs,
      prior = lda_prior
    ),
    random_forest = run_fold_rf(
      fold_inputs,
      ntree = rf_ntree
    ),
    decision_tree = run_fold_dt(
      fold_inputs,
      maxdepth = dt_maxdepth
    )
  )

  for (model_name in model_names) {
    model_out <- model_outputs[[model_name]]

    cv_predictions[[model_name]][[fold]] <- data.frame(
      ID = cv_test$ID,
      matched_set = cv_test$matched_set,
      observed = cv_test$CC_bin,
      prob = model_out$prob,
      model = model_name,
      fold = fold
    )

    feature_tracker[[model_name]][[fold]] <- model_out$selected_features
    tuning_tracker[[model_name]][[fold]] <- model_out$tuning
    feature_count_tracker[[model_name]][fold] <- length(model_out$selected_features)
  }
}

cv_summary <- bind_rows(lapply(model_names, function(model_name) {
  pred_df <- bind_rows(cv_predictions[[model_name]])

  roc_obj_cv <- roc(pred_df$observed, pred_df$prob, quiet = TRUE)

  data.frame(
    model = model_name,
    n_predictions = nrow(pred_df),
    cv_auc = as.numeric(auc(roc_obj_cv)),
    cv_accuracy = mean(ifelse(pred_df$prob > 0.5, 1, 0) == pred_df$observed)
  )
}))

print(cv_summary %>% arrange(desc(cv_auc)))
write.csv(cv_summary,
          "Tables/AUC and Accuracy Per Model.csv",
          row.names = FALSE)

```

```{r, echo=FALSE, results='asis'}

render_html_table(
  cv_summary %>% arrange(desc(cv_auc)),
  caption = "Cross-validation AUC and accuracy by model",
  digits = 3
)

```


### Stable Features in CV Folds
Summarize feature selection stability across folds for each model

```{r, echo=FALSE, results='asis'}

for (model_name in model_names) {
  display_name <- model_display_name(model_name)
  feature_counts <- sort(table(unlist(feature_tracker[[model_name]])), decreasing = TRUE)

  cat("#### ", display_name, "\n\n", sep = "")

  if (length(feature_counts) == 0) {
    cat("<p>No features were selected for this model in any fold.</p>\n")
  } else {
    stable_features <- data.frame(
      miRNA = names(head(feature_counts, 20)),
      n_folds_selected = as.integer(head(feature_counts, 20)),
      row.names = NULL
    )
    render_html_table(
      stable_features,
      caption = paste("Top stable features for", display_name)
    )
  }

  fold_feature_counts <- data.frame(
    fold = seq_along(feature_count_tracker[[model_name]]),
    n_selected_features = feature_count_tracker[[model_name]],
    row.names = NULL
  )
  render_html_table(
    fold_feature_counts,
    caption = paste("Selected feature counts by fold for", display_name)
  )

  tuning_summary <- bind_rows(tuning_tracker[[model_name]])
  render_html_table(
    tuning_summary,
    caption = paste("Tuning summary for", display_name),
    digits = 3
  )
}

fold_filter_summary <- data.frame(
  fold = seq_along(filter_sizes),
  n_selected_miRNA = filter_sizes,
  row.names = NULL
)

render_html_table(
  fold_filter_summary,
  caption = paste("Number of", fold_filter_name, "selected miRNAs by fold")
)

```

## Training/Holdout Evaluation

```{r, echo=FALSE, results='asis'}

filter_out_final <- fold_filter_fn(
  train_fold_dat = train_set,
  miRNA_cols = miRNA_cols,
  top_n = fold_filter_top_n
)
final_fold_inputs <- prepare_fold_inputs(
  train_set,
  holdout,
  filter_out_final$selected
)

cat(
  "<p>Number of final ",
  fold_filter_name,
  " selected miRNAs: ",
  length(filter_out_final$selected),
  "</p>\n",
  sep = ""
)

final_model_outputs <- list(
  forward_step = run_fold_forward_step(
    final_fold_inputs,
    max_steps = forward_step_max_steps
  ),
  lasso = run_fold_lasso(
    final_fold_inputs,
    alpha_lasso = lasso_alpha
  ),
  enet = run_fold_enet(
    final_fold_inputs,
    alpha_enet = enet_alpha_grid
  ),
  lda = run_fold_lda(
    final_fold_inputs,
    prior = lda_prior
  ),
  random_forest = run_fold_rf(
    final_fold_inputs,
    ntree = rf_ntree
  ),
  decision_tree = run_fold_dt(
    final_fold_inputs,
    maxdepth = dt_maxdepth
  )
)

holdout_predictions <- bind_rows(lapply(model_names, function(model_name) {
  data.frame(
    model = model_name,
    observed = holdout$CC_bin,
    prob = as.numeric(final_model_outputs[[model_name]]$prob),
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(
    predicted_class = if_else(.data$prob > 0.5, 1, 0)
  )

holdout_summary <- holdout_predictions %>%
  group_by(.data$model) %>%
  summarise(
    test_accuracy = mean(.data$predicted_class == .data$observed),
    test_auc = as.numeric(pROC::auc(pROC::roc(.data$observed, .data$prob, quiet = TRUE))),
    .groups = "drop"
  )

write.csv(holdout_summary,
          "Tables/Holdout Summary.csv",
          row.names = FALSE)

render_html_table(
  holdout_summary %>% arrange(desc(.data$test_auc)),
  caption = "Holdout AUC and accuracy by model",
  digits = 3
)

final_model_diagnostics <- bind_rows(lapply(model_names, function(model_name) {
  model_out <- final_model_outputs[[model_name]]

  model_out$tuning %>%
    mutate(
      n_prob = length(model_out$prob),
      .before = 1
    )
}))

diagnostic_summary <- final_model_diagnostics %>%
  select(tidyselect::all_of(c("model", "n_prob", "n_selected_features")))

render_html_table(
  diagnostic_summary,
  caption = "Final model diagnostics"
)

holdout_feature_summary <- final_model_diagnostics %>%
  select(tidyselect::all_of(c("model", "n_selected_features")))

render_html_table(
  holdout_feature_summary,
  caption = "Selected feature counts in final holdout models"
)

```

## Per-miRNA Logistic Regression Analysis

```{r, echo=FALSE, results='asis'}

interpretation_fn <- function(
  train_dat,
  selected_miRNA,
  model_formula = CC_bin ~ miRNA_value + Sex + Race + Age
) {
  if (!length(selected_miRNA)) {
    stop("No selected miRNAs were provided.", call. = TRUE)
  }

# we know that there are no missing demographic variables 

  metadata_train <- data.frame(
    ID = as.character(train_dat$ID),
    CC = factor(train_dat$CC, levels = c("Controls", "PDAC case")),
    CC_bin = ifelse(train_dat$CC == "PDAC case", 1, 0),
    Sex = factor(train_dat$Sex, levels = c("Female", "Male")),
    Race = factor(train_dat$Race, levels = c("Asian", "Black", "White", "Others")),
    Age = as.numeric(train_dat$Age),
    row.names = NULL
  )

  complete_rows <- complete.cases(metadata_train)
  metadata_train <- droplevels(metadata_train[complete_rows, , drop = FALSE])

  if (nrow(metadata_train) == 0) {
    stop("No complete cases remained after screening the interpretation dataset metadata.", call. = TRUE)
  }

  feature_mat <- train_dat[complete_rows, selected_miRNA, drop = FALSE]

  selected_miRNA <- selected_miRNA[
    colSums(is.na(feature_mat)) == 0
  ]

  if (!length(selected_miRNA)) {
    stop("All selected miRNAs were removed after screening for complete data in the interpretation dataset.", call. = TRUE)
  }

  result_rows <- lapply(selected_miRNA, function(miRNA_name) {
    fit_dat <- metadata_train
    fit_dat$miRNA_value <- feature_mat[[miRNA_name]]

    glm_fit <- stats::glm(model_formula, data = fit_dat, family = stats::binomial())

    coef_table <- summary(glm_fit)$coefficients

    if (!("miRNA_value" %in% rownames(coef_table))) {
      stop(sprintf("Fitted model for %s did not return a miRNA_value coefficient.", miRNA_name), call. = TRUE)
    }

    estimate <- coef_table["miRNA_value", "Estimate"]
    std_error <- coef_table["miRNA_value", "Std. Error"]
    z_value <- coef_table["miRNA_value", "z value"]
    p_value <- coef_table["miRNA_value", "Pr(>|z|)"]

    data.frame(
      miRNA = miRNA_name,
      estimate = estimate,
      odds_ratio = exp(estimate),
      std_error = std_error,
      z_value = z_value,
      p_value = p_value,
      row.names = NULL
    )
  })

  association_results <- do.call(rbind, result_rows)
  association_results <- association_results[stats::complete.cases(association_results$p_value), , drop = FALSE]

  if (nrow(association_results) == 0) {
    stop("No usable per-miRNA logistic regression results were produced.", call. = TRUE)
  }

  association_results$adj_p_value <- stats::p.adjust(association_results$p_value, method = "BH")
  association_results <- association_results %>%
    arrange(desc(.data$adj_p_value), .data$p_value)

  list(
    results = association_results,
    metadata = metadata_train
  )
}

final_filter_stats <- filter_out_final$stats
association_out <- interpretation_fn(
  train_dat = train_set,
  selected_miRNA = filter_out_final$selected,
  model_formula = CC_bin ~ miRNA_value + Sex + Race + Age
)

association_results <- association_out$results

write.csv(association_results,
          "Tables/Per-miRNA Logistic Regression Results.csv",
          row.names = FALSE)

render_html_table(
  association_results,
  caption = "Per-miRNA logistic regression results",
  digits = 3
)

```

## Calibration Curves and Additional Metrics
Reliability diagrams (calibration curves) and Brier scores for each model on the holdout set

```{r, echo=FALSE, results='asis', fig.width=7, fig.height=5}

calibration_summary <- holdout_predictions %>%
  group_by(.data$model) %>%
  summarise(
    brier_score = mean((.data$prob - .data$observed)^2, na.rm = TRUE),
    .groups = "drop"
  )

calibration_plot_data <- holdout_predictions %>%
  mutate(
    prob_bin = cut(
      .data$prob,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  group_by(.data$model, .data$prob_bin) %>%
  summarise(
    mean_prob = mean(.data$prob, na.rm = TRUE),
    observed_rate = mean(.data$observed, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(.data$mean_prob), !is.na(.data$observed_rate))

calibration_plots <- list()

for (model_name in model_names) {
  calibration_data <- calibration_plot_data %>%
    filter(.data$model == model_name)
  brier_score <- calibration_summary %>%
    filter(.data$model == model_name) %>%
    pull("brier_score")

  calibration_plot <- ggplot(calibration_data, aes(x = mean_prob, y = observed_rate)) +
    geom_point(aes(size = n)) +
    geom_line() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(
      title = paste("Calibration Curve -", model_name),
      subtitle = paste("Brier score =", round(brier_score, 4)),
      x = "Mean Predicted Probability",
      y = "Observed Event Rate",
      size = "Bin N"
    ) +
    theme_minimal()

  calibration_plots[[model_name]] <- calibration_plot

  ggsave(
    filename = paste0("calibration_curve_", model_name, ".png"),
    plot = calibration_plot,
    width = 7,
    height = 5,
    units = "in",
    dpi = 72
  )
}

calibration_summary <- calibration_summary %>%
  arrange(.data$brier_score)

write.csv(calibration_summary, "Tables/Calibration Summary.csv", row.names = FALSE)

render_html_table(
  calibration_summary,
  caption = "Brier scores by model",
  digits = 4
)

for (model_name in model_names) {
  cat("### ", model_display_name(model_name), "\n\n", sep = "")
  print(calibration_plots[[model_name]])
}

```


# Interpretation and Visualization
ROC curve comparison of all models on the test set.

```{r, echo=FALSE, results='asis', fig.width=7, fig.height=5}

# ROC curve plot with all six models on the holdout set
render_html_table(
  holdout_summary %>% arrange(desc(.data$test_auc)),
  caption = "Holdout ROC performance summary",
  digits = 3
)

roc_curves <- lapply(model_names, function(model_name) {
  roc(holdout$CC_bin, final_model_outputs[[model_name]]$prob, quiet = TRUE)
})
names(roc_curves) <- model_names

roc_labels <- vapply(model_names, function(model_name) {
  sprintf("%s (AUC = %.3f)", model_name, as.numeric(auc(roc_curves[[model_name]])))
}, character(1))

roc_label_order <- roc_labels[holdout_summary$model[order(-holdout_summary$test_auc)]]

roc_curve_data <- bind_rows(lapply(model_names, function(model_name) {
  roc_obj <- roc_curves[[model_name]]

  data.frame(
    false_positive_rate = 1 - roc_obj$specificities,
    true_positive_rate = roc_obj$sensitivities,
    legend_label = roc_labels[[model_name]],
    row.names = NULL
  )
}))

roc_curve_data$legend_label <- factor(roc_curve_data$legend_label, levels = roc_label_order)

roc_curve_plot <- ggplot(
  roc_curve_data,
  aes(x = false_positive_rate, y = true_positive_rate, color = legend_label)
) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "ROC Curve Comparison of All Models",
    x = "FP Rate",
    y = "TP Rate",
    color = "Model"
  ) +
  theme_minimal()

ggsave(
  filename = "roc_curve_comparison.png",
  plot = roc_curve_plot,
  width = 7,
  height = 5,
  units = "in",
  dpi = 72
)

print(roc_curve_plot)


```

# Selected miRNA Features and Interpretation

```{r, echo=FALSE, results='asis'}

# final feature counts across all models:
# create a table of all miRNA features selected by any final model, with a count of how many models selected each feature, sorted by count and then alphabetically
final_selected_features <- bind_rows(lapply(model_names, function(model_name) {
  model_selected <- final_model_outputs[[model_name]]$selected_features

  if (is.null(model_selected) || !length(model_selected)) {
    return(data.frame(
      model = character(0),
      miRNA = character(0),
      stringsAsFactors = FALSE
    ))
  }

  data.frame(
    model = rep(model_name, length(model_selected)),
    miRNA = model_selected,
    stringsAsFactors = FALSE
  )
}))

final_model_feature_counts <- final_selected_features %>%
  distinct(model, miRNA) %>%
  count(miRNA, name = "n_models_selected") %>%
  arrange(desc(n_models_selected), miRNA)

write.csv(final_model_feature_counts, "Tables/Final Model Feature Counts.csv", row.names = FALSE)

render_html_table(
  final_model_feature_counts,
  caption = "Counts of models selecting each miRNA"
)

if (nrow(final_model_feature_counts) == 0) {
  stop("No miRNAs were selected by any final model.", call. = TRUE)
}

top_ranked_miRNA <- final_model_feature_counts %>%
  slice_max(order_by = n_models_selected, n = 1, with_ties = TRUE) %>%
  arrange(miRNA)
top_rank_count <- top_ranked_miRNA$n_models_selected[[1]]

cat(
  "<p>Top-rank selection count: ",
  top_rank_count,
  " of ",
  length(model_names),
  " models</p>\n",
  sep = ""
)

render_html_table(
  top_ranked_miRNA,
  caption = "Top-ranked consensus miRNAs"
)

```

# Final Summary Table of Consensus Features
A table of all miRNAs selected by any final model, with their filter statistics, logistic regression results, model selection counts, and annotation names.

```{r, echo=FALSE, results='asis'}

# table of consensus features, with logistic-regression results,
# model selection counts, filter stats, and annotation names

gpl <- read_tsv(
  "GSE314577_family.xml/GPL30934-tbl-1.txt",
  col_names = c("Annotation", "Alias", "MIMAT", "ClassName"),
  show_col_types = FALSE
)

final_feature_table <- data.frame(miRNA = top_ranked_miRNA$miRNA, row.names = NULL) %>%
  left_join(final_filter_stats[, filter_stat_cols], by = "miRNA") %>%
  left_join(association_results[, association_cols], by = "miRNA") %>%
  left_join(final_model_feature_counts[, c("miRNA", "n_models_selected")], by = "miRNA") %>%
  left_join(select(gpl, miRNA = MIMAT, Annotation), by = "miRNA") %>%
  arrange(adj_p_value)

write.csv(final_feature_table, "Tables/Final Consensus Features.csv", row.names = FALSE)

render_html_table(
  final_feature_table,
  caption = "Consensus feature summary",
  digits = 3
)

```

## Association Plot of Final Consensus Features
Logistic-regression association plot of final consensus features, colored by class SNR
```{r, echo=FALSE, fig.width=7, fig.height=5}

association_plot <- ggplot(
  final_feature_table,
  aes(x = estimate, y = -log10(adj_p_value), color = class_snr)
) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red", na.value = "grey") +
  labs(
    title = "Association Plot of Final Consensus Features",
    x = "Log-Odds Coefficient",
    y = "-Log10 Adjusted P-Value",
    color = "Class SNR"
  ) +
  theme_minimal()

print(association_plot)

ggsave(
  filename = "Figures/Association Plot Final Features.png",
  plot = association_plot,
  width = 7,
  height = 5,
  units = "in",
  dpi = 72
)

```
