---
title: "Final Project - "
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

setwd("~/Classes/BMI7235 - Machine Learning/Final Project")

library(readxl)
library(readr)
library(tidyr)
library(glmnet)
library(ggplot2)
library(pROC)
library(dplyr)
library(rpart)
library(randomForest)

```

```{r helpers, include=FALSE}

# Prepare the training and evaluation datasets for model fitting based on the selected miRNAs from the fold-specific feature filter. This includes creating the model matrices and formulas needed for each model runner function.
prepare_fold_inputs <- function(train_fold_dat, valid_fold_dat, selected_miRNA) {
  train_fold_dat$CC_bin <- ifelse(train_fold_dat$CC == "PDAC case", 1, 0)
  valid_fold_dat$CC_bin <- ifelse(valid_fold_dat$CC == "PDAC case", 1, 0)

  if (!length(selected_miRNA)) {
    stop("prepare_fold_inputs failed: feature filtering returned zero miRNAs.", call. = FALSE)
  }

  complete_feature_mask <- 
    colSums(is.na(train_fold_dat[, selected_miRNA, drop = FALSE])) == 0 &
      colSums(is.na(valid_fold_dat[, selected_miRNA, drop = FALSE])) == 0
  selected_miRNA <- selected_miRNA[complete_feature_mask]

  if (!length(selected_miRNA)) {
    stop("prepare_fold_inputs failed: all selected miRNAs were removed after screening for complete data in both training and evaluation sets.", call. = FALSE)
  }

  keep_cols <- c("CC", "CC_bin", selected_miRNA)
  train_model_dat <- train_fold_dat[, keep_cols, drop = FALSE]
  valid_model_dat <- valid_fold_dat[, keep_cols, drop = FALSE]
  model_formula <- stats::reformulate(selected_miRNA, response = "CC_bin")
  class_formula <- stats::reformulate(selected_miRNA, response = "CC")

  list(
    selected_miRNA = selected_miRNA,
    model_formula = model_formula,
    class_formula = class_formula,
    x_train = stats::model.matrix(model_formula, data = train_model_dat)[, -1, drop = FALSE],
    x_valid = stats::model.matrix(model_formula, data = valid_model_dat)[, -1, drop = FALSE],
    y_train = train_model_dat$CC_bin,
    train_model_dat = train_model_dat,
    valid_model_dat = valid_model_dat
  )
}

# Feature filter
#  The top N miRNAs are selected based on the highest SNR values, with ties broken by p-value from t-test
filter_fold_class_snr <- function(train_fold_dat, miRNA_cols, top_n = 50) {
  control_rows <- train_fold_dat$CC == "Controls"
  case_rows <- train_fold_dat$CC == "PDAC case"

  if (!any(control_rows) || !any(case_rows)) {
    stop("filter_fold_class_snr failed: the training fold does not contain both control and PDAC case samples.", call. = FALSE)
  }

  control_mat <- train_fold_dat[control_rows, miRNA_cols, drop = FALSE]
  case_mat <- train_fold_dat[case_rows, miRNA_cols, drop = FALSE]

  class_stats <- data.frame(
    miRNA = miRNA_cols,
    mean_control = colMeans(control_mat, na.rm = TRUE),
    mean_case = colMeans(case_mat, na.rm = TRUE),
    sd_control = apply(control_mat, 2, sd, na.rm = TRUE),
    sd_case = apply(case_mat, 2, sd, na.rm = TRUE),
    row.names = NULL
  )

  class_stats$mean_diff <- class_stats$mean_case - class_stats$mean_control
  class_stats$class_snr <- with(
    class_stats,
    ifelse(sd_case + sd_control > 0, abs(mean_diff) / (sd_case + sd_control), NA_real_)
  )

  t_test_stats <- t(vapply(
    miRNA_cols,
    function(miRNA) {
      test_out <- try(
        stats::t.test(case_mat[[miRNA]], control_mat[[miRNA]]),
        silent = TRUE
      )

      if (inherits(test_out, "try-error")) {
        return(c(t_stat = NA_real_, p_value = NA_real_))
      }

      c(
        t_stat = unname(test_out$statistic),
        p_value = test_out$p.value
      )
    },
    numeric(2)
  ))

  class_stats$t_stat <- t_test_stats[, "t_stat"]
  class_stats$p_value <- t_test_stats[, "p_value"]

  class_stats <- class_stats[
    stats::complete.cases(class_stats[, c("class_snr", "p_value")]),
    ,
    drop = FALSE
  ]

  if (nrow(class_stats) == 0) {
    stop("filter_fold_class_snr failed: no miRNAs had usable class SNR and p-value statistics after filtering.", call. = FALSE)
  }

  class_stats <- class_stats[order(-class_stats$class_snr, class_stats$p_value), , drop = FALSE]

  selected <- head(class_stats$miRNA, top_n)

  if (!length(selected)) {
    stop("filter_fold_class_snr failed: ranking produced zero selected miRNAs.", call. = FALSE)
  }

  list(
    stats = class_stats,
    selected = selected
  )
}

run_fold_class_snr <- function(train_fold_dat, miRNA_cols, top_n = 50) {
  # Ensure binary outcome is coded consistently
  train_fold_dat$CC <- factor(train_fold_dat$CC, levels = c("Controls", "PDAC case"))

  # Compute class-specific summary statistics for each miRNA
  class_stats <- data.frame(
    miRNA = miRNA_cols,
    mean_case = sapply(train_fold_dat[, miRNA_cols, drop = FALSE], function(x) {
      mean(x[train_fold_dat$CC == "PDAC case"], na.rm = TRUE)
    }),
    mean_control = sapply(train_fold_dat[, miRNA_cols, drop = FALSE], function(x) {
      mean(x[train_fold_dat$CC == "Controls"], na.rm = TRUE)
    }),
    sd_case = sapply(train_fold_dat[, miRNA_cols, drop = FALSE], function(x) {
      sd(x[train_fold_dat$CC == "PDAC case"], na.rm = TRUE)
    }),
    sd_control = sapply(train_fold_dat[, miRNA_cols, drop = FALSE], function(x) {
      sd(x[train_fold_dat$CC == "Controls"], na.rm = TRUE)
    }),
    row.names = NULL
  )

  # Difference in means
  class_stats$mean_diff <- class_stats$mean_case - class_stats$mean_control

  # Classic two-class signal-to-noise ratio
  class_stats$class_snr <- with(
    class_stats,
    ifelse(
      (sd_case + sd_control) > 0,
      abs(mean_diff) / (sd_case + sd_control),
      NA_real_
    )
  )

  # Secondary ranking statistic: ordinary two-sample t-test p-value
  class_stats$p_value <- sapply(class_stats$miRNA, function(m) {
    x <- train_fold_dat[[m]]
    tryCatch(
      t.test(x ~ train_fold_dat$CC)$p.value,
      error = function(e) NA_real_
    )
  })

  # Clean and rank
  class_stats <- class_stats[is.finite(class_stats$class_snr) & is.finite(class_stats$p_value), ]
  class_stats <- class_stats[order(-class_stats$class_snr, class_stats$p_value), ]

  selected <- head(class_stats$miRNA, top_n)

  list(
    stats = class_stats,
    selected = selected
  )
}
```

```{r constants, include=FALSE}
# constants and model parameters

#random seed
set.seed(31415)

model_names <- c("forward_step", "lasso", "enet", "lda", "random_forest", "decision_tree")
enet_alpha_grid <- c(0.1, 0.3, 0.5, 0.7, 0.9)
forward_step_max_steps <- 20
lasso_alpha <- 1
lda_prior <- c(0.5, 0.5)
rf_ntree <- 1000
dt_maxdepth <- 4

fold_filter_name <- "class SNR"
fold_filter_top_n <- 50
interpretation_method_name <- "per-miRNA logistic regression on PDAC status"
filter_stat_cols <- c("miRNA", "mean_control", "mean_case", "mean_diff", "sd_control", "sd_case", "class_snr", "p_value")
association_cols <- c("miRNA", "estimate", "odds_ratio", "std_error", "z_value", "p_value", "adj_p_value")

```


# Preparing Data

## Import Data
We use the normalized miRNA count matrix from GEO and merge it with the demographic metadata.

```{r}

normalized_counts <- read_xlsx("GSE314577/GSE314577_VUMC_miRNA_normalized_data.xlsx")
metadata <- read_xlsx("GSE314577/GSE314577_VUMC_Metadata_ch_revised.xlsx")

normalized_counts <- normalized_counts %>%
  mutate(ID = as.character(.data$ID))

metadata <- metadata %>%
  mutate(
    ID = as.character(.data$ID),
    CC = factor(.data$CC, levels = c("Controls", "PDAC case")),
    Sex = factor(.data$Sex, levels = c("Female", "Male")),
    Race = factor(.data$Race, levels = c("Asian", "Black", "White", "Others")),
    matched_set = factor(.data$matched_set),
    Age = as.numeric(.data$age_sam),
    CC_bin = ifelse(.data$CC == "PDAC case", 1, 0)
  )

full_dat <- metadata %>%
  left_join(normalized_counts, by = "ID")

miRNA_cols <- grep("^MIMAT", names(full_dat), value = TRUE)

cat("Rows/columns in merged dataset:", dim(full_dat), "\n")
cat("Number of miRNA columns:", length(miRNA_cols), "\n")

```

## Create Training and Holdout Sets
To properly split the data while respecting the matched-pairs, we will perform a training/holdout split at the level of matched sets, ensuring that all samples from a given matched set are either in the training set or holdout set, but not both.

```{r}

all_sets <- unique(full_dat$matched_set)
n_train_sets <- floor(0.8 * length(all_sets))
train_set_ids <- sample(all_sets, n_train_sets)

train_set <- full_dat %>% filter(matched_set %in% train_set_ids)
holdout  <- full_dat %>% filter(!matched_set %in% train_set_ids)

cat("Training rows:", nrow(train_set), "\n")
cat("Holdout rows:", nrow(holdout), "\n")
cat("Training class balance:\n")
print(table(train_set$CC))
cat("Holdout class balance:\n")
print(table(holdout$CC))

```

# Cross-Validation Loop
We perform a k-fold cross-validation on the training set of paired samples, ensuring matched-samples are kept together in the same fold.

Within each fold:
1. We apply a fold-specific feature filter, currently class-based signal-to-noise ratio (SNR) on the normalized counts, to select the top miRNAs within the training fold.
2. We then fit each model using only the selected miRNAs from the normalized count matrix.

The SNR filtering is calculated as the absolute difference in means between cases and controls divided by the sum of standard deviations, with an additional t-test for association. The top N miRNAs are selected based on the highest SNR values, with ties broken by p-value.

```{r, include=FALSE}

# Model Runners
run_fold_forward_step <- function(fold_inputs, max_steps = 10) {
  null_glm <- glm(CC_bin ~ 1, data = fold_inputs$train_model_dat, family = binomial())
  full_glm <- glm(fold_inputs$model_formula, data = fold_inputs$train_model_dat, family = binomial())

  forward_glm <- step(
    object = null_glm,
    scope = list(lower = formula(null_glm), upper = formula(full_glm)),
    direction = "forward",
    trace = 0,
    steps = min(max_steps, length(fold_inputs$selected_miRNA))
  )

  prob <- as.numeric(
    predict(forward_glm, newdata = fold_inputs$valid_model_dat, type = "response")
  )

  selected_features <- attr(terms(forward_glm), "term.labels")

  list(
    prob = prob,
    selected_features = selected_features,
    tuning = data.frame(
      model = "forward_step",
      n_selected_features = length(selected_features),
      n_terms = length(selected_features),
      stringsAsFactors = FALSE
    )
  )
}

run_fold_lasso <- function(fold_inputs, alpha_lasso = 1) {
  cv.lasso <- glmnet::cv.glmnet(
    fold_inputs$x_train,
    fold_inputs$y_train,
    alpha = alpha_lasso,
    family = "binomial"
  )

  bestlam_lasso <- cv.lasso$lambda.min
  prob <- as.numeric(
    predict(cv.lasso, s = bestlam_lasso, newx = fold_inputs$x_valid, type = "response")
  )

  coef_mat_lasso <- as.matrix(
    predict(cv.lasso, type = "coefficients", s = bestlam_lasso)
  )

  selected_features <- rownames(coef_mat_lasso)[coef_mat_lasso[, 1] != 0]
  selected_features <- setdiff(selected_features, "(Intercept)")

  list(
    prob = prob,
    selected_features = selected_features,
    tuning = data.frame(
      model = "lasso",
      n_selected_features = length(selected_features),
      alpha = alpha_lasso,
      lambda = bestlam_lasso,
      stringsAsFactors = FALSE
    )
  )
}

run_fold_enet <- function(
  fold_inputs,
  alpha_enet = c(0.1, 0.3, 0.5, 0.7, 0.9)
) {
  alpha_grid <- sort(unique(alpha_enet))
  cv_fits <- lapply(alpha_grid, function(alpha_value) {
    glmnet::cv.glmnet(
      fold_inputs$x_train,
      fold_inputs$y_train,
      alpha = alpha_value,
      family = "binomial"
    )
  })
  cv_scores <- vapply(cv_fits, function(cv_fit) min(cv_fit$cvm), numeric(1))
  best_index <- which.min(cv_scores)

  cv.enet <- cv_fits[[best_index]]
  best_alpha_enet <- alpha_grid[[best_index]]
  bestlam_enet <- cv.enet$lambda.min
  prob <- as.numeric(
    predict(cv.enet, s = bestlam_enet, newx = fold_inputs$x_valid, type = "response")
  )

  coef_mat_enet <- as.matrix(
    predict(cv.enet, type = "coefficients", s = bestlam_enet)
  )

  selected_features <- rownames(coef_mat_enet)[coef_mat_enet[, 1] != 0]
  selected_features <- setdiff(selected_features, "(Intercept)")

  list(
    prob = prob,
    selected_features = selected_features,
    tuning = data.frame(
      model = "enet",
      n_selected_features = length(selected_features),
      alpha = best_alpha_enet,
      lambda = bestlam_enet,
      cv_score = cv_scores[[best_index]],
      alpha_grid = paste(alpha_grid, collapse = ","),
      stringsAsFactors = FALSE
    )
  )
}

run_fold_lda <- function(
  fold_inputs,
  max_features = 20,
  prior = c(0.5, 0.5)
) {
  selected_miRNA <- head(fold_inputs$selected_miRNA, max_features)
  class_formula <- stats::reformulate(selected_miRNA, response = "CC")
  train_model_dat <- fold_inputs$train_model_dat[, c("CC", "CC_bin", selected_miRNA), drop = FALSE]
  valid_model_dat <- fold_inputs$valid_model_dat[, c("CC", "CC_bin", selected_miRNA), drop = FALSE]

  lda_fit <- MASS::lda(
    class_formula,
    data = train_model_dat,
    prior = prior
  )
  lda_pred <- predict(lda_fit, newdata = valid_model_dat)
  prob <- as.numeric(lda_pred$posterior[, "PDAC case"])

  list(
    prob = prob,
    selected_features = selected_miRNA,
    tuning = data.frame(
      model = "lda",
      n_selected_features = length(selected_miRNA),
      n_features = length(selected_miRNA),
      prior = paste(prior, collapse = ","),
      stringsAsFactors = FALSE
    )
  )
}

run_fold_rf <- function(fold_inputs, ntree = 500) {
  rf_fit <- randomForest::randomForest(
    fold_inputs$class_formula,
    data = fold_inputs$train_model_dat,
    ntree = ntree
  )

  prob <- as.numeric(
    predict(rf_fit, newdata = fold_inputs$valid_model_dat, type = "prob")[, "PDAC case"]
  )

  list(
    prob = prob,
    selected_features = fold_inputs$selected_miRNA,
    tuning = data.frame(
      model = "random_forest",
      n_selected_features = length(fold_inputs$selected_miRNA),
      ntree = ntree,
      stringsAsFactors = FALSE
    )
  )
}

run_fold_dt <- function(fold_inputs, cp_dt = 0.01, maxdepth = 10) {
  dt_fit <- rpart::rpart(
    fold_inputs$class_formula,
    data = fold_inputs$train_model_dat,
    method = "class",
    control = rpart::rpart.control(cp = cp_dt, maxdepth = maxdepth)
  )

  prob <- as.numeric(
    predict(dt_fit, newdata = fold_inputs$valid_model_dat, type = "prob")[, "PDAC case"]
  )

  selected_features <- attr(terms(dt_fit), "term.labels")

  list(
    prob = prob,
    selected_features = selected_features,
    tuning = data.frame(
      model = "decision_tree",
      n_selected_features = length(selected_features),
      cp = cp_dt,
      maxdepth = maxdepth,
      n_terms = length(selected_features),
      stringsAsFactors = FALSE
    )
  )
}

```
```{r}

fold_filter_fn <- run_fold_class_snr

cv_set_ids <- unique(train_set$matched_set)
k <- 5
fold_id <- sample(rep(1:k, length.out = length(cv_set_ids)))

fold_map <- data.frame(
  matched_set = cv_set_ids,
  fold = fold_id
)

cv_predictions <- list()
feature_tracker <- list()
tuning_tracker <- list()
feature_count_tracker <- list()
filter_sizes <- integer(k)

for (model_name in model_names) {
  cv_predictions[[model_name]] <- list()
  feature_tracker[[model_name]] <- list()
  tuning_tracker[[model_name]] <- list()
  feature_count_tracker[[model_name]] <- integer(k)
}

for (fold in 1:k) {
  cat("Running fold", fold, "of", k, "...\n")

  test_sets <- fold_map$matched_set[fold_map$fold == fold]

  cv_train <- train_set %>% filter(!matched_set %in% test_sets)
  cv_test <- train_set %>% filter(matched_set %in% test_sets)

  filter_out <- fold_filter_fn(
    train_fold_dat = cv_train,
    miRNA_cols = miRNA_cols,
    top_n = fold_filter_top_n
  )
  fold_inputs <- prepare_fold_inputs(
    cv_train,
    cv_test,
    filter_out$selected
  )

  filter_sizes[fold] <- length(filter_out$selected)

  model_outputs <- list(
    forward_step = run_fold_forward_step(
      fold_inputs,
      max_steps = forward_step_max_steps
    ),
    lasso = run_fold_lasso(
      fold_inputs,
      alpha_lasso = lasso_alpha
    ),
    enet = run_fold_enet(
      fold_inputs,
      alpha_enet = enet_alpha_grid
    ),
    lda = run_fold_lda(
      fold_inputs,
      prior = lda_prior
    ),
    random_forest = run_fold_rf(
      fold_inputs,
      ntree = rf_ntree
    ),
    decision_tree = run_fold_dt(
      fold_inputs,
      maxdepth = dt_maxdepth
    )
  )

  for (model_name in model_names) {
    model_out <- model_outputs[[model_name]]

    cv_predictions[[model_name]][[fold]] <- data.frame(
      ID = cv_test$ID,
      matched_set = cv_test$matched_set,
      observed = cv_test$CC_bin,
      prob = model_out$prob,
      model = model_name,
      fold = fold
    )

    feature_tracker[[model_name]][[fold]] <- model_out$selected_features
    tuning_tracker[[model_name]][[fold]] <- model_out$tuning
    feature_count_tracker[[model_name]][fold] <- length(model_out$selected_features)
  }
}

cv_summary <- bind_rows(lapply(model_names, function(model_name) {
  pred_df <- bind_rows(cv_predictions[[model_name]])

  roc_obj_cv <- roc(pred_df$observed, pred_df$prob, quiet = TRUE)

  data.frame(
    model = model_name,
    n_predictions = nrow(pred_df),
    cv_auc = as.numeric(auc(roc_obj_cv)),
    cv_accuracy = mean(ifelse(pred_df$prob > 0.5, 1, 0) == pred_df$observed)
  )
}))

print(cv_summary[order(-cv_summary$cv_auc), ])
write.csv(cv_summary,
          "Tables/AUC and Accuracy Per Model.csv",
          row.names = FALSE)

```


## Stable Features in CV Folds
Summarize feature selection stability across folds for each model

```{r}

for (model_name in model_names) {
  feature_counts <- sort(table(unlist(feature_tracker[[model_name]])), decreasing = TRUE)

  cat("Top stable features for", model_name, ":\n")
  if (length(feature_counts) == 0) {
    cat("No features were selected for this model in any fold.\n")
  } else {
    print(head(feature_counts, 20))
  }

  cat("Number of selected features by fold for", model_name, ":\n")
  print(feature_count_tracker[[model_name]])

  cat("Tuning summary for", model_name, ":\n")
  print(bind_rows(tuning_tracker[[model_name]]))
}

cat("Number of", fold_filter_name, "-selected miRNAs by fold:\n")
print(filter_sizes)

```

## Training/Holdout Evaluation

```{r}

filter_out_final <- fold_filter_fn(
  train_fold_dat = train_set,
  miRNA_cols = miRNA_cols,
  top_n = fold_filter_top_n
)
final_fold_inputs <- prepare_fold_inputs(
  train_set,
  holdout,
  filter_out_final$selected
)

cat("Number of final", fold_filter_name, "-selected miRNAs:", length(filter_out_final$selected), "\n")

final_model_outputs <- list(
  forward_step = run_fold_forward_step(
    final_fold_inputs,
    max_steps = forward_step_max_steps
  ),
  lasso = run_fold_lasso(
    final_fold_inputs,
    alpha_lasso = lasso_alpha
  ),
  enet = run_fold_enet(
    final_fold_inputs,
    alpha_enet = enet_alpha_grid
  ),
  lda = run_fold_lda(
    final_fold_inputs,
    prior = lda_prior
  ),
  random_forest = run_fold_rf(
    final_fold_inputs,
    ntree = rf_ntree
  ),
  decision_tree = run_fold_dt(
    final_fold_inputs,
    maxdepth = dt_maxdepth
  )
)

holdout_summary <- bind_rows(lapply(model_names, function(model_name) {
  model_out <- final_model_outputs[[model_name]]
  holdout_class <- ifelse(model_out$prob > 0.5, 1, 0)
  roc_obj_test <- roc(holdout$CC_bin, model_out$prob, quiet = TRUE)

  data.frame(
    model = model_name,
    test_accuracy = mean(holdout_class == holdout$CC_bin),
    test_auc = as.numeric(auc(roc_obj_test))
  )
}))

print(holdout_summary[order(-holdout_summary$test_auc), ])
write.csv(holdout_summary,
          "Tables/Holdout Summary.csv",
          row.names = FALSE)

lapply(final_model_outputs, function(x) {
  list(
    n_prob = length(x$prob),
    n_selected = if (!is.null(x$selected_features)) length(x$selected_features) else NA,
    tuning = x$tuning
  )
})

holdout_feature_summary <- bind_rows(lapply(model_names, function(model_name) {
  model_out <- final_model_outputs[[model_name]]
  data.frame(
    model = model_name,
    n_selected_features = if (!is.null(model_out$selected_features)) length(model_out$selected_features) else NA
  )
}))

print(holdout_feature_summary)

```

## Per-miRNA Logistic Regression Analysis

```{r}

run_feature_logistic <- function(
  train_dat,
  selected_miRNA,
  model_formula = CC_bin ~ miRNA_value + Sex + Race + Age
) {
  if (!length(selected_miRNA)) {
    stop("run_feature_logistic failed: no selected miRNAs were provided.", call. = FALSE)
  }

  metadata_train <- data.frame(
    ID = as.character(train_dat$ID),
    CC = factor(train_dat$CC, levels = c("Controls", "PDAC case")),
    CC_bin = ifelse(train_dat$CC == "PDAC case", 1, 0),
    Sex = factor(train_dat$Sex, levels = c("Female", "Male")),
    Race = factor(train_dat$Race, levels = c("Asian", "Black", "White", "Others")),
    Age = as.numeric(train_dat$Age),
    row.names = NULL
  )

  complete_rows <- complete.cases(metadata_train)
  metadata_train <- droplevels(metadata_train[complete_rows, , drop = FALSE])

  if (nrow(metadata_train) == 0) {
    stop("run_feature_logistic failed: no complete cases remained after screening the interpretation dataset metadata.", call. = FALSE)
  }

  feature_mat <- train_dat[complete_rows, selected_miRNA, drop = FALSE]

  selected_miRNA <- selected_miRNA[
    colSums(is.na(feature_mat)) == 0
  ]

  if (!length(selected_miRNA)) {
    stop("run_feature_logistic failed: all selected miRNAs were removed after screening for complete data in the interpretation dataset.", call. = FALSE)
  }

  result_rows <- lapply(selected_miRNA, function(miRNA_name) {
    fit_dat <- metadata_train
    fit_dat$miRNA_value <- feature_mat[[miRNA_name]]

    glm_fit <- try(
      suppressWarnings(
        stats::glm(model_formula, data = fit_dat, family = stats::binomial())
      ),
      silent = TRUE
    )

    if (inherits(glm_fit, "try-error")) {
      stop(sprintf("run_feature_logistic failed: logistic regression did not converge for %s.", miRNA_name), call. = FALSE)
    }

    coef_table <- summary(glm_fit)$coefficients

    if (!("miRNA_value" %in% rownames(coef_table))) {
      stop(sprintf("run_feature_logistic failed: fitted model for %s did not return a miRNA_value coefficient.", miRNA_name), call. = FALSE)
    }

    estimate <- coef_table["miRNA_value", "Estimate"]
    std_error <- coef_table["miRNA_value", "Std. Error"]
    z_value <- coef_table["miRNA_value", "z value"]
    p_value <- coef_table["miRNA_value", "Pr(>|z|)"]

    data.frame(
      miRNA = miRNA_name,
      estimate = estimate,
      odds_ratio = exp(estimate),
      std_error = std_error,
      z_value = z_value,
      p_value = p_value,
      row.names = NULL
    )
  })

  association_results <- do.call(rbind, result_rows)
  association_results <- association_results[stats::complete.cases(association_results$p_value), , drop = FALSE]

  if (nrow(association_results) == 0) {
    stop("run_feature_logistic failed: no usable per-miRNA logistic regression results were produced.", call. = FALSE)
  }

  association_results$adj_p_value <- stats::p.adjust(association_results$p_value, method = "BH")
  association_results <- association_results[
    order(association_results$adj_p_value, association_results$p_value),
    ,
    drop = FALSE
  ]

  list(
    results = association_results,
    metadata = metadata_train
  )
}

interpretation_fn <- run_feature_logistic
final_filter_stats <- filter_out_final$stats
association_out <- interpretation_fn(
  train_dat = train_set,
  selected_miRNA = filter_out_final$selected,
  model_formula = CC_bin ~ miRNA_value + Sex + Race + Age
)

association_results <- association_out$results

print(head(association_results))
write.csv(association_results,
          "Tables/Per-miRNA Logistic Regression Results.csv",
          row.names = FALSE)

```

# Calibration Curves and Additional Metrics
Reliability diagrams (calibration curves) and Brier scores for each model on the holdout set

```{r}

calibration_summary <- data.frame(
  model = character(),
  brier_score = numeric(),
  stringsAsFactors = FALSE
)

for (model_name in model_names) {
  model_out <- final_model_outputs[[model_name]]
  calibration_raw <- data.frame(
    observed = holdout$CC_bin,
    prob = as.numeric(model_out$prob)
  )

  brier_score <- mean((calibration_raw$prob - calibration_raw$observed)^2, na.rm = TRUE)
  calibration_summary <- bind_rows(
  calibration_summary,
  data.frame(
    model = model_name,
    brier_score = brier_score
  )
)
  # Bin probabilities for reliability curve
  calibration_data <- calibration_raw %>%
    mutate(
      prob_bin = cut(
        prob,
        breaks = seq(0, 1, by = 0.1),
        include.lowest = TRUE,
        right = TRUE
      )
    ) %>%
    group_by(prob_bin) %>%
    summarise(
      mean_prob = mean(prob, na.rm = TRUE),
      observed_rate = mean(observed, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    filter(!is.na(mean_prob), !is.na(observed_rate))
  # Save calibration plot
  png(
    paste0("calibration_curve_", model_name, ".png"),
    width = 7 * 72,
    height = 5 * 72
  )
  print(
    ggplot(calibration_data, aes(x = mean_prob, y = observed_rate)) +
      geom_point(aes(size = n)) +
      geom_line() +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
      coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
      labs(
        title = paste("Calibration Curve -", model_name),
        subtitle = paste("Brier score =", round(brier_score, 4)),
        x = "Mean Predicted Probability",
        y = "Observed Event Rate",
        size = "Bin N"
      ) +
      theme_minimal()
  )
  invisible(dev.off())
}

# Show Brier scores for all models
calibration_summary <- calibration_summary %>%
  arrange(brier_score)
print(calibration_summary)

write.csv(calibration_summary, "Tables/Calibration Summary.csv", row.names = FALSE)

```


# Interpretation and Visualization
ROC curve comparison of all models on the test set.

```{r}

# ROC curve plot with all six models on the holdout set
print(holdout_summary[order(-holdout_summary$test_auc), ])

roc_curves <- lapply(model_names, function(model_name) {
  roc(holdout$CC_bin, final_model_outputs[[model_name]]$prob, quiet = TRUE)
})
names(roc_curves) <- model_names

roc_labels <- vapply(model_names, function(model_name) {
  sprintf("%s (AUC = %.3f)", model_name, as.numeric(auc(roc_curves[[model_name]])))
}, character(1))

roc_label_order <- roc_labels[holdout_summary$model[order(-holdout_summary$test_auc)]]

roc_curve_data <- bind_rows(lapply(model_names, function(model_name) {
  roc_obj <- roc_curves[[model_name]]

  data.frame(
    false_positive_rate = 1 - roc_obj$specificities,
    true_positive_rate = roc_obj$sensitivities,
    legend_label = roc_labels[[model_name]],
    row.names = NULL
  )
}))

roc_curve_data$legend_label <- factor(roc_curve_data$legend_label, levels = roc_label_order)

roc_curve_plot <- ggplot(
  roc_curve_data,
  aes(x = false_positive_rate, y = true_positive_rate, color = legend_label)
) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "ROC Curve Comparison of All Models",
    x = "False Positive Rate",
    y = "True Positive Rate",
    color = "Model"
  ) +
  theme_minimal()

print(roc_curve_plot)
png("roc_curve_comparison.png", width = 7 * 72, height = 5 * 72)
print(roc_curve_plot)
invisible(dev.off())


```

# Selected miRNA Features and Interpretation

```{r}

# final feature counts across all models:
# create a table of all miRNA features selected by any final model, with a count of how many models selected each feature, sorted by count and then alphabetically
final_model_feature_counts <- data.frame(
  miRNA = miRNA_cols,
  n_models_selected = integer(length(miRNA_cols)),
  row.names = NULL
)

for (model_name in model_names) {
  model_selected <- final_model_outputs[[model_name]]$selected_features

  if (!is.null(model_selected) && length(model_selected) > 0) {
    final_model_feature_counts$n_models_selected[
      final_model_feature_counts$miRNA %in% model_selected
    ] <- final_model_feature_counts$n_models_selected[
      final_model_feature_counts$miRNA %in% model_selected
    ] + 1L
  }
}

final_model_feature_counts <- final_model_feature_counts[
  final_model_feature_counts$n_models_selected > 0,
  ,
  drop = FALSE
]
final_model_feature_counts <- final_model_feature_counts[
  order(-final_model_feature_counts$n_models_selected, final_model_feature_counts$miRNA),
  ,
  drop = FALSE
]

print(final_model_feature_counts)
write.csv(final_model_feature_counts, "Tables/Final Model Feature Counts.csv", row.names = FALSE)

if (nrow(final_model_feature_counts) == 0) {
  stop("final consensus feature selection failed: no miRNAs were selected by any final model.", call. = FALSE)
}

top_rank_count <- max(final_model_feature_counts$n_models_selected)
top_ranked_miRNA <- final_model_feature_counts[
  final_model_feature_counts$n_models_selected == top_rank_count,
  ,
  drop = FALSE
]

cat("Top-rank selection count:", top_rank_count, "of", length(model_names), "models\n")
cat("Top-ranked miRNAs:", paste(top_ranked_miRNA$miRNA, collapse = ", "), "\n")
print(top_ranked_miRNA)

```

# Final Summary Table of Consensus Features
A table of all miRNAs selected by any final model, with their filter statistics, logistic regression results, model selection counts, and annotation names.

```{r}

# table of consensus features, with logistic-regression results,
# model selection counts, filter stats, and annotation names

gpl <- read_tsv(
  "GSE314577_family.xml/GPL30934-tbl-1.txt",
  col_names = c("Annotation", "Alias", "MIMAT", "ClassName"),
  show_col_types = FALSE
)

final_feature_table <- data.frame(miRNA = top_ranked_miRNA$miRNA, row.names = NULL) %>%
  left_join(final_filter_stats[, filter_stat_cols], by = "miRNA") %>%
  left_join(association_results[, association_cols], by = "miRNA") %>%
  left_join(final_model_feature_counts[, c("miRNA", "n_models_selected")], by = "miRNA") %>%
  left_join(select(gpl, miRNA = MIMAT, Annotation), by = "miRNA") %>%
  arrange(adj_p_value)

print(final_feature_table)

write.csv(final_feature_table, "Tables/Final Consensus Features.csv", row.names = FALSE)

```


## Association Plot of Final Consensus Features
Logistic-regression association plot of final consensus features, colored by class SNR
```{r}

png("Figures/Association Plot Final Features.png", width = 7 * 72, height = 5 * 72)
ggplot(final_feature_table, aes(x = estimate, y = -log10(adj_p_value), color = class_snr)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red", na.value = "grey") +
  labs(title = "Association Plot of Final Consensus Features", x = "Log-Odds Coefficient", y = "-Log10 Adjusted P-Value", color = "Class SNR") +
  theme_minimal()
invisible(dev.off())

```
