---
title: "Final Project - Predicting PDAC with Circulating miRNA"
authors: "Madison Blake, Kristyn Gumpper-Fedus, Sean Kirkpatrick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  dev = "png",
  fig.path = "Figures/",
  fig.retina = 1
)

# Create output directories for figures and tables
dir.create("Figures", showWarnings = FALSE, recursive = TRUE)
dir.create("Tables", showWarnings = FALSE, recursive = TRUE)
dir.create("Data", showWarnings = FALSE, recursive = TRUE)

library(readxl)
library(readr)
library(tidyr)
library(glmnet)
library(ggplot2)
library(pROC)
library(dplyr)
library(rpart)
library(randomForest)
library(limma)
library(multiMiR)
library(mice)

source("R/plotHelpers.R")
source("R/run_fold_lasso.R")
source("R/run_fold_enet.R")
source("R/run_fold_forward_step.R")
source("R/run_fold_lda.R")
source("R/run_fold_rf.R")
source("R/run_fold_dt.R")
source("R/run_limma_paired.R")
source("R/differential_expression_filter_fn.R")
source("R/multi_mir_analysis.R")

```

```{r runtime_tracking_start, include=FALSE}

report_start_time <- Sys.time()

```

```{r report_table_styles, echo=FALSE, results='asis'}

cat(
  paste(
    "<style>",
    "table {",
    "  border-collapse: collapse;",
    "}",
    "table th,",
    "table td {",
    "  border: 1px solid #cfcfcf;",
    "  padding: 4px 8px;",
    "}",
    "</style>",
    sep = "\n"
  )
)

```

```{r constants, include=FALSE}
# constants and model parameters

#random seed
set.seed(31415)

#necessary external data files
normalized_data_file <- "Data/GSE314577/GSE314577_VUMC_miRNA_normalized_data.xlsx"
metadata_file <- "Data/GSE314577/GSE314577_VUMC_Metadata_ch_revised.xlsx"
gene_family_file <- "Data/GSE314577_family.xml/GPL30934-tbl-1.txt"

#model parameters and settings
model_names <- c("forward_step", "lasso", "enet", "lda", "random_forest", "decision_tree")
enet_alpha_grid <- c(0.5, 0.6, 0.75, 0.9, 0.95)
forward_step_max_steps <- 25
lasso_alpha <- 1
lda_prior <- c(0.5, 0.5)
rf_ntree <- 300
dt_maxdepth <- 7

log2_pseudocount <- 1
run_limma_analysis <- TRUE

fold_filter_name <- "blocked limma"
fold_filter_top_n <- 20
final_variable_heatmap_max_n <- fold_filter_top_n

run_multimir_analysis <- TRUE
multimir_results_rds_file <- "Data/Top-ranked consensus miRNA multiMiR results.rds"
multimir_shared_gene_thresholds <- c(2L, 3L, 4L)
multimir_pathway_enrichment_min_shared_miRNA <- 3L
multimir_pathway_enrichment_top_terms <- 15L

demographic_covariates <- c("Sex", "Race", "Age")
factor_demographic_covariates <- c("Sex", "Race")
use_model_covariates <- TRUE

limma_group_term <- "CC"
limma_adjustment_covariates <- demographic_covariates
filter_stat_cols <- c(
  "miRNA",
  "mean_control",
  "mean_case",
  "mean_diff",
  "limma_log_fc",
  "limma_avg_expr",
  "limma_t",
  "limma_p_value",
  "limma_adj_p_value",
  "limma_b"
)
```

```{r write_comma_separated_list, include=FALSE}

write_comma_separated_list <- function(items) {
  if (length(items) == 0) {
    return("")
  } else if (length(items) == 1) {
    return(items)
  } else {
    return(paste(
      paste(items[-length(items)], collapse = ", "),
      items[length(items)],
      sep = ifelse(length(items) > 2, ", and ", " and ")
    ))
  }
}

```

```{r dev_block, include=FALSE}

##### DEVELOPMENT #####
developmentMode <- FALSE
development_use_model_covariates <- FALSE

# models to exclude
if (developmentMode) {
  # development_excluded_models <- c("random_forest", "decision_tree")
  development_excluded_models <- c()

  use_model_covariates <- development_use_model_covariates
  model_names <- setdiff(model_names, development_excluded_models)
}

model_covariates <- if (use_model_covariates) demographic_covariates else character(0)
model_covariate_text <- if (length(model_covariates)) {
  write_comma_separated_list(model_covariates)
} else {
  "no demographic covariates"
}
model_design_text <- if (length(model_covariates)) {
  paste("together with", model_covariate_text, "in the design")
} else {
  "without additional demographic covariates in the predictive design"
}
limma_adjustment_covariate_text <- write_comma_separated_list(limma_adjustment_covariates)

```

```{r model_dispatch_helpers, include=FALSE}

model_runner_map <- list(
  forward_step = function(fold_inputs) {
    run_fold_forward_step(
      fold_inputs,
      max_steps = forward_step_max_steps
    )
  },
  lasso = function(fold_inputs) {
    run_fold_lasso(
      fold_inputs,
      alpha_lasso = lasso_alpha
    )
  },
  enet = function(fold_inputs) {
    run_fold_enet(
      fold_inputs,
      alpha_enet = enet_alpha_grid
    )
  },
  lda = function(fold_inputs) {
    run_fold_lda(
      fold_inputs,
      prior = lda_prior
    )
  },
  random_forest = function(fold_inputs) {
    run_fold_rf(
      fold_inputs,
      ntree = rf_ntree
    )
  },
  decision_tree = function(fold_inputs) {
    run_fold_dt(
      fold_inputs,
      maxdepth = dt_maxdepth
    )
  }
)

run_active_models <- function(fold_inputs, active_model_names = model_names) {
  unknown_model_names <- setdiff(active_model_names, names(model_runner_map))

  if (length(unknown_model_names)) {
    stop(
      sprintf(
        "Requested models are not registered: %s",
        write_comma_separated_list(unknown_model_names)
      ),
      call. = TRUE
    )
  }

  model_outputs <- lapply(active_model_names, function(model_name) {
    model_runner_map[[model_name]](fold_inputs)
  })
  names(model_outputs) <- active_model_names

  model_outputs
}

```

# Preparing Data

## Import Data
We use the normalized miRNA count matrix from GEO, apply a single log2 transform with a pseudocount of `r log2_pseudocount`, and then merge it with the demographic metadata. Categorical variables are set to factors and the PDAC case and control variable is set to binary for modeling.

That imported log-transformed miRNA matrix is then used throughout the analysis, including blocked limma and all downstream predictive modeling.

```{r data_import, include=FALSE}

normalized_counts <- read_xlsx(normalized_data_file)
metadata <- read_xlsx(metadata_file)

raw_miRNA_values <- as.numeric(as.matrix(normalized_counts[, grep("^MIMAT", names(normalized_counts))]))


normalized_counts <- normalized_counts %>%
  mutate(ID = as.character(.data$ID)) %>%
  mutate(across(starts_with("MIMAT"), ~ log2(.x + log2_pseudocount)))

transformed_miRNA_values <- as.numeric(as.matrix(normalized_counts[, grep("^MIMAT", names(normalized_counts))]))

metadata <- metadata %>%
  mutate(
    ID = as.character(.data$ID),
    CC = factor(.data$CC, levels = c("Controls", "PDAC case")),
    Sex = factor(.data$Sex, levels = c("Female", "Male")),
    Race = factor(.data$Race, levels = c("Asian", "Black", "White", "Others")),
    matched_set = factor(.data$matched_set),
    Age = as.numeric(.data$age_sam),
    CC_bin = ifelse(.data$CC == "PDAC case", 1, 0)
  )

full_dat <- metadata %>%
  left_join(normalized_counts, by = "ID")

miRNA_all_columns <- grep("^MIMAT", names(full_dat), value = TRUE)

cat("Rows/columns in merged dataset:", dim(full_dat), "\n")
cat("Number of miRNA columns:", length(miRNA_all_columns), "\n")

```

The side-by-side QQ plots below show how the global miRNA value distribution shifts after the import-time log2 transform.

```{r log2_transform_qq_comparison, echo=FALSE, fig.width=10, fig.height=5}

plot_settings <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
on.exit(par(plot_settings), add = TRUE)

qqnorm(
  raw_miRNA_values,
  main = "Before Log2 Transform",
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles",
  pch = 16,
  cex = 0.3,
  col = rgb(59 / 255, 91 / 255, 146 / 255, 0.25)
)
qqline(raw_miRNA_values, col = "#C65D3B", lwd = 2)

qqnorm(
  transformed_miRNA_values,
  main = paste0("After Log2(x + ", log2_pseudocount, ")"),
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles",
  pch = 16,
  cex = 0.3,
  col = rgb(59 / 255, 91 / 255, 146 / 255, 0.25)
)
qqline(transformed_miRNA_values, col = "#C65D3B", lwd = 2)

```

```{r missing_value_check, include=FALSE}

md.pattern(full_dat, plot = FALSE)
# the full data set is fully observed

```
The dataset is fully observed with no missing values, so we can proceed without imputation.

## Create Training and Holdout Sets
To split the data while respecting matched-pairs, we performed the split on the matched set IDs, ensuring each pair is not split between sets.

```{r training_split, echo=FALSE, results='asis'}

all_sets <- unique(full_dat$matched_set)
n_train_sets <- floor(0.8 * length(all_sets))
train_set_ids <- sample(all_sets, n_train_sets)

train_set <- full_dat %>% filter(matched_set %in% train_set_ids)
holdout  <- full_dat %>% filter(!matched_set %in% train_set_ids)

split_summary <- data.frame(
  dataset = c("Training", "Holdout"),
  rows = c(nrow(train_set), nrow(holdout)),
  row.names = NULL
)

class_balance <- bind_rows(
  data.frame(
    dataset = "Training",
    class = names(table(train_set$CC)),
    n_samples = as.integer(table(train_set$CC)),
    row.names = NULL
  ),
  data.frame(
    dataset = "Holdout",
    class = names(table(holdout$CC)),
    n_samples = as.integer(table(holdout$CC)),
    row.names = NULL
  )
)

render_html_table(split_summary, caption = "Training and holdout set sizes")
render_html_table(class_balance, caption = "Training and holdout class balance")

```

## Perform Differential Expression Filtering
The limma filtering is based on a blocked linear model with duplicate-correlation adjustment for matched sets and is run once on the full training set. Because the miRNA columns are log2-transformed during import, blocked limma and the downstream predictive models operate on the same log-expression scale. In that model, `r limma_group_term` is the term of interest and `r limma_adjustment_covariate_text` are adjustment covariates. The top N miRNAs are selected based on the smallest limma p-values, with ties broken by the absolute moderated t-statistic. To avoid rerunning the expensive blocked-limma fit on every render, the report loads the saved full-training limma object by default; set `run_limma_analysis <- TRUE` once after changing upstream preprocessing if you need to rebuild `Data/Full train_filter_results.rds`.
```{r limma_filter_results, eval=run_limma_analysis, include=FALSE}
train_filter_results <- differential_expression_filter_fn(
  train_fold_dat = train_set,
  miRNA_cols = miRNA_all_columns,
  filter_stat_cols = filter_stat_cols,
  top_n = fold_filter_top_n,
  group_term = limma_group_term,
  adjustment_covariates = limma_adjustment_covariates
)
```

```{r store_limma_results, eval=run_limma_analysis, include=FALSE}

# store the full train_filter_results object to disk
saveRDS(train_filter_results, "Data/Full train_filter_results.rds")

```

```{r load_filter_results_from_disk, eval=!run_limma_analysis, include=FALSE}

train_filter_results <- readRDS("Data/Full train_filter_results.rds")

```

```{r limma_filter_summary, echo=FALSE, results='asis'}

write.csv(
  train_filter_results$stats %>% arrange(limma_p_value),
  "Tables/Blocked limma filter results on full training set.csv",
  row.names = FALSE
)

```

## Top `r fold_filter_top_n` miRNAs selected by the blocked limma filter on the full training set
```{r top_limma_miRNA_table, echo=FALSE, results='asis'}

# train_filter_results$selected is the vector of miRNAs passed downstream for modeling.
# Here we show the top-ranked miRNAs by p-value with their limma statistics for context.

top_ranked_miRNA <- train_filter_results$stats %>%
  slice_min(order_by = limma_p_value, n = fold_filter_top_n, with_ties = TRUE) %>%
  arrange(limma_p_value, desc(abs(limma_t)), miRNA)

render_html_table(
  top_ranked_miRNA,
  caption = paste(
    "Top",
    fold_filter_top_n,
    fold_filter_name,
    "selected miRNAs on the full training set"
  ),
  digits = 4
)

write.csv(
  top_ranked_miRNA,
  paste0("Tables/Top ", fold_filter_top_n, " ", fold_filter_name, " selected miRNAs with stats on full training set.csv"),
  row.names = FALSE
)

write.csv(
  train_filter_results$selected,
  paste0("Tables/Limma selected miRNAs on full training set.csv"),
  row.names = FALSE
)

#describe shape of top_ranked_miRNA table and train_filter_results$selected vector for context
cat(
  "The top-ranked miRNAs table has",
  nrow(top_ranked_miRNA),
  "rows and",
  ncol(top_ranked_miRNA),
  "columns. The selected miRNAs vector has",
  length(train_filter_results$selected),
  "miRNAs."
)

#compare the different lists of miRNA and build a table showing the overlap between the two rankings
overlap_table <- data.frame(
  miRNA = train_filter_results$selected,
  in_top_ranked = train_filter_results$selected %in% top_ranked_miRNA$miRNA
)

```
These top `r fold_filter_top_n` miRNAs are then used as the shared candidate feature set for all models in the subsequent cross-validation step, with the rationale that this filtering step is necessary to reduce the feature space to a level that can be effectively handled by the models given the sample size, while still being broad enough to allow for model-specific feature selection and not biasing towards any one model's selection criteria.

# Cross-Validation Loop
```{r k_definition, include=FALSE}

k <- 5

```
We perform a `r k`-fold cross-validation on the training set of paired samples, ensuring matched-samples are kept together in the same fold.

Within each fold:
1. We restrict the log2-transformed normalized count matrix to the top `r fold_filter_top_n` miRNAs selected from the limma analysis on the full training set.
2. We then fit each model using that shared subset of selected miRNAs `r model_design_text`.
```{r fold_input_helpers, include=FALSE}

map_to_training_levels <- function(values, training_levels) {
  if (!length(training_levels)) {
    stop("No training levels were available for factor alignment.", call. = TRUE)
  }

  mapped_values <- as.character(values)
  mapped_values[!(mapped_values %in% training_levels)] <- training_levels[[1]]
  factor(mapped_values, levels = training_levels)
}

build_penalty_factor <- function(column_names, protected_terms) {
  penalty_factor <- rep(1, length(column_names))

  for (term_name in protected_terms) {
    penalty_factor[startsWith(column_names, term_name)] <- 0
  }

  penalty_factor
}

is_model_covariate_term <- function(variable_names, covariate_names = model_covariates) {
  if (!length(variable_names)) {
    return(logical(0))
  }

  if (!length(covariate_names)) {
    return(rep(FALSE, length(variable_names)))
  }

  vapply(
    variable_names,
    function(variable_name) any(startsWith(variable_name, covariate_names)),
    logical(1)
  )
}

prepare_fold_inputs <- function(train_fold_dat, valid_fold_dat, selected_miRNA) {
  train_fold_dat <- train_fold_dat %>%
    mutate(CC_bin = if_else(.data$CC == "PDAC case", 1, 0))
  valid_fold_dat <- valid_fold_dat %>%
    mutate(CC_bin = if_else(.data$CC == "PDAC case", 1, 0))

  if (!length(selected_miRNA)) {
    stop("Feature filtering returned zero miRNAs.", call. = TRUE)
  }

  keep_cols <- c("CC", "CC_bin", model_covariates, selected_miRNA)
  train_model_dat <- train_fold_dat %>%
    dplyr::select(tidyselect::all_of(keep_cols)) %>%
    droplevels()
  valid_model_dat <- valid_fold_dat %>%
    dplyr::select(tidyselect::all_of(keep_cols))

  factor_covariates <- intersect(factor_demographic_covariates, model_covariates)

  for (covariate_name in factor_covariates) {
    valid_model_dat[[covariate_name]] <- map_to_training_levels(
      valid_model_dat[[covariate_name]],
      levels(train_model_dat[[covariate_name]])
    )
  }

  raw_model_formula <- stats::reformulate(c(model_covariates, selected_miRNA), response = "CC_bin")
  train_design_matrix <- stats::model.matrix(raw_model_formula, data = train_model_dat)
  valid_design_matrix <- stats::model.matrix(raw_model_formula, data = valid_model_dat)
  x_train <- train_design_matrix[, -1, drop = FALSE]
  x_valid <- valid_design_matrix[, -1, drop = FALSE]
  design_feature_cols <- colnames(x_train)
  base_feature_cols <- design_feature_cols[!(design_feature_cols %in% selected_miRNA)]
  model_formula <- stats::reformulate(design_feature_cols, response = "CC_bin")
  base_model_formula <- if (length(base_feature_cols)) {
    stats::reformulate(base_feature_cols, response = "CC_bin")
  } else {
    stats::as.formula("CC_bin ~ 1")
  }
  class_formula <- stats::reformulate(design_feature_cols, response = "CC")
  train_design_dat <- data.frame(
    CC = train_model_dat$CC,
    CC_bin = train_model_dat$CC_bin,
    as.data.frame(x_train, check.names = FALSE),
    check.names = FALSE
  )
  valid_design_dat <- data.frame(
    CC = valid_model_dat$CC,
    CC_bin = valid_model_dat$CC_bin,
    as.data.frame(x_valid, check.names = FALSE),
    check.names = FALSE
  )

  list(
    selected_miRNA = selected_miRNA,
    model_covariates = model_covariates,
    design_feature_cols = design_feature_cols,
    base_feature_cols = base_feature_cols,
    base_model_formula = base_model_formula,
    model_formula = model_formula,
    class_formula = class_formula,
    x_train = x_train,
    x_valid = x_valid,
    penalty_factor = build_penalty_factor(design_feature_cols, model_covariates),
    y_train = train_model_dat$CC_bin,
    train_model_dat = train_model_dat,
    valid_model_dat = valid_model_dat,
    train_design_dat = train_design_dat,
    valid_design_dat = valid_design_dat
  )
}
```

```{r cross_validation_run, include=FALSE}

cv_set_ids <- unique(train_set$matched_set)
fold_id <- sample(rep(1:k, length.out = length(cv_set_ids)))

fold_map <- data.frame(
  matched_set = cv_set_ids,
  fold = fold_id
)

cv_predictions <- list()
feature_tracker <- list()
tuning_tracker <- list()
feature_count_tracker <- list()
filter_sizes <- integer(k)

for (model_name in model_names) {
  cv_predictions[[model_name]] <- list()
  feature_tracker[[model_name]] <- list()
  tuning_tracker[[model_name]] <- list()
  feature_count_tracker[[model_name]] <- integer(k)
}

for (fold in 1:k) {
  cat("Running fold", fold, "of", k, "...\n")

  test_sets <- fold_map$matched_set[fold_map$fold == fold]

  cv_train <- train_set %>% filter(!matched_set %in% test_sets)
  cv_test <- train_set %>% filter(matched_set %in% test_sets)

  fold_inputs <- prepare_fold_inputs(
    cv_train,
    cv_test,
    top_ranked_miRNA$miRNA
    # train_filter_results$selected
  )

  filter_sizes[fold] <- length(fold_inputs$selected_miRNA)

  model_outputs <- run_active_models(fold_inputs)

  for (model_name in model_names) {
    model_out <- model_outputs[[model_name]]

    cv_predictions[[model_name]][[fold]] <- data.frame(
      ID = cv_test$ID,
      matched_set = cv_test$matched_set,
      observed = cv_test$CC_bin,
      prob = model_out$prob,
      model = model_name,
      fold = fold
    )

    feature_tracker[[model_name]][[fold]] <- model_out$selected_miRNA
    tuning_tracker[[model_name]][[fold]] <- model_out$tuning
    feature_count_tracker[[model_name]][fold] <- length(model_out$selected_miRNA)
  }
}

cv_summary <- bind_rows(lapply(model_names, function(model_name) {
  pred_df <- bind_rows(cv_predictions[[model_name]])

  roc_obj_cv <- roc(pred_df$observed, pred_df$prob, quiet = TRUE)

  data.frame(
    model = model_name,
    n_predictions = nrow(pred_df),
    cv_auc = as.numeric(auc(roc_obj_cv)),
    cv_accuracy = mean(ifelse(pred_df$prob > 0.5, 1, 0) == pred_df$observed)
  )
}))

print(cv_summary %>% arrange(desc(cv_auc)))
write.csv(cv_summary,
          "Tables/AUC and Accuracy Per Model.csv",
          row.names = FALSE)

```

```{r cv_summary_table, echo=FALSE, results='asis'}

render_html_table(
  cv_summary %>% arrange(desc(cv_auc)),
  caption = "Cross-validation AUC and accuracy by model",
  digits = 3
)

```


### Stable Features in CV Folds
Summarize feature selection stability across folds for each model

```{r stable_features_table, echo=FALSE, results='asis'}

for (model_name in model_names) {
  display_name <- model_display_name(model_name)
  feature_counts <- sort(table(unlist(feature_tracker[[model_name]])), decreasing = TRUE)

  cat("#### ", display_name, "\n\n", sep = "")

  if (length(feature_counts) == 0) {
    cat("<p>No features were selected for this model in any fold.</p>\n")
  } else {
    stable_features <- data.frame(
      miRNA = names(head(feature_counts, 20)),
      n_folds_selected = as.integer(head(feature_counts, 20)),
      row.names = NULL
    )
    render_html_table(
      stable_features,
      caption = paste("Top stable features for", display_name)
    )
  }

  fold_feature_counts <- data.frame(
    fold = seq_along(feature_count_tracker[[model_name]]),
    n_selected_miRNA = feature_count_tracker[[model_name]],
    row.names = NULL
  )
  render_html_table(
    fold_feature_counts,
    caption = paste("Selected miRNA counts by fold for", display_name)
  )

  tuning_summary <- bind_rows(tuning_tracker[[model_name]])
  render_html_table(
    tuning_summary,
    caption = paste("Tuning summary for", display_name),
    digits = 3
  )
}

fold_filter_summary <- data.frame(
  fold = seq_along(filter_sizes),
  n_selected_miRNA = filter_sizes,
  row.names = NULL
)

render_html_table(
  fold_filter_summary,
  caption = paste("Number of", fold_filter_name, "preselected miRNAs available by fold")
)

```

## Holdout Evaluation
Evaluate the final fold, using the calibrated models from the CV step, on the holdout set. We report AUC and accuracy for each model on the holdout set, as well as the number of features selected in each final model and the number of miRNAs that passed the initial limma filter.
```{r holdout_evaluation, include=FALSE}

final_filter_results <- train_filter_results
final_fold_inputs <- prepare_fold_inputs(
  train_set,
  holdout,
  top_ranked_miRNA$miRNA
  # train_filter_results$selected
)

cat(
  "<p>Number of final ",
  fold_filter_name,
  " selected miRNAs: ",
  length(top_ranked_miRNA$miRNA),
  "</p>\n",
  sep = ""
)

final_model_outputs <- run_active_models(final_fold_inputs)
```

```{r holdout_predictions, echo=FALSE, results='asis'}

holdout_predictions <- bind_rows(lapply(model_names, function(model_name) {
  data.frame(
    model = model_name,
    observed = final_fold_inputs$valid_model_dat$CC_bin,
    prob = as.numeric(final_model_outputs[[model_name]]$prob),
    stringsAsFactors = FALSE
  )
})) %>%
  mutate(
    predicted_class = if_else(.data$prob > 0.5, 1, 0)
  )

holdout_summary <- holdout_predictions %>%
  group_by(.data$model) %>%
  summarise(
    test_accuracy = mean(.data$predicted_class == .data$observed),
    test_auc = as.numeric(pROC::auc(pROC::roc(.data$observed, .data$prob, quiet = TRUE))),
    .groups = "drop"
  )

write.csv(holdout_summary,
          "Tables/Holdout Summary.csv",
          row.names = FALSE)

render_html_table(
  holdout_summary %>% arrange(desc(.data$test_auc)),
  caption = "Holdout AUC and accuracy by model",
  digits = 3
)

final_model_diagnostics <- bind_rows(lapply(model_names, function(model_name) {
  model_out <- final_model_outputs[[model_name]]

  model_out$tuning %>%
    mutate(
      n_prob = length(model_out$prob),
      .before = 1
    )
}))

diagnostic_summary <- final_model_diagnostics %>%
  dplyr::select(tidyselect::all_of(c("model", "n_prob", "n_selected_miRNA")))

render_html_table(
  diagnostic_summary,
  caption = "Final model diagnostics"
)

holdout_feature_summary <- final_model_diagnostics %>%
  dplyr::select(tidyselect::all_of(c("model", "n_selected_miRNA")))

render_html_table(
  holdout_feature_summary,
  caption = "Selected miRNA counts in final holdout models"
)

final_model_variable_summary <- bind_rows(lapply(model_names, function(model_name) {
  model_out <- final_model_outputs[[model_name]]
  selected_variables <- unique(model_out$selected_variables)
  selected_demographics <- selected_variables[is_model_covariate_term(selected_variables)]
  selected_miRNA <- selected_variables[!is_model_covariate_term(selected_variables)]

  data.frame(
    model = model_display_name(model_name),
    n_final_variables = length(selected_variables),
    demographic_variables = if (length(selected_demographics)) {
      write_comma_separated_list(selected_demographics)
    } else {
      "None"
    },
    miRNA_variables = if (length(selected_miRNA)) {
      write_comma_separated_list(selected_miRNA)
    } else {
      "None"
    },
    final_variables = if (length(selected_variables)) {
      write_comma_separated_list(selected_variables)
    } else {
      "None"
    },
    stringsAsFactors = FALSE
  )
}))

write.csv(
  final_model_variable_summary,
  "Tables/Final Model Variables.csv",
  row.names = FALSE
)

render_html_table(
  final_model_variable_summary,
  caption = "Final variables used by each holdout model"
)

final_model_variable_long <- bind_rows(lapply(model_names, function(model_name) {
  selected_variables <- unique(final_model_outputs[[model_name]]$selected_variables)

  if (!length(selected_variables)) {
    return(data.frame(
      model = character(0),
      variable = character(0),
      variable_type = character(0),
      stringsAsFactors = FALSE
    ))
  }

  data.frame(
    model = model_display_name(model_name),
    variable = selected_variables,
    variable_type = ifelse(
      is_model_covariate_term(selected_variables),
      "Demographic",
      "miRNA"
    ),
    stringsAsFactors = FALSE
  )
}))

final_variable_presence <- final_model_variable_long %>%
  distinct(model, variable, .keep_all = TRUE) %>%
  mutate(selected = "Yes") %>%
  tidyr::pivot_wider(
    names_from = model,
    values_from = selected,
    values_fill = ""
  ) %>%
  left_join(
    final_model_variable_long %>%
      distinct(model, variable, .keep_all = TRUE) %>%
      count(variable, variable_type, name = "n_models_selected"),
    by = c("variable", "variable_type")
  ) %>%
  mutate(
    variable_type_order = if_else(.data$variable_type == "Demographic", 0L, 1L)
  ) %>%
  arrange(.data$variable_type_order, desc(.data$n_models_selected), .data$variable) %>%
  dplyr::select(-.data$variable_type_order)

write.csv(
  final_variable_presence,
  "Tables/Final Variable Inclusion Matrix.csv",
  row.names = FALSE
)

render_html_table(
  final_variable_presence,
  caption = "Final variable inclusion matrix across holdout models"
)

final_model_variable_heatmap_counts <- final_model_variable_long %>%
  distinct(model, variable, .keep_all = TRUE) %>%
  count(variable, variable_type, name = "n_models_selected") %>%
  arrange(desc(n_models_selected), variable_type, variable)

if (!is.null(final_variable_heatmap_max_n) &&
    (!is.numeric(final_variable_heatmap_max_n) || length(final_variable_heatmap_max_n) != 1L || final_variable_heatmap_max_n < 1)) {
  stop("final_variable_heatmap_max_n must be NULL or a single positive number.")
}

final_variable_heatmap_top_n <- if (is.null(final_variable_heatmap_max_n)) {
  nrow(final_model_variable_heatmap_counts)
} else {
  min(as.integer(final_variable_heatmap_max_n), nrow(final_model_variable_heatmap_counts))
}

final_variable_heatmap_subtitle <- if (final_variable_heatmap_top_n < nrow(final_model_variable_heatmap_counts)) {
  paste(
    "Top",
    final_variable_heatmap_top_n,
    "variables ordered by number of final models selecting them"
  )
} else {
  "All final variables ordered by number of final models selecting them"
}

final_model_display_names <- vapply(model_names, model_display_name, character(1))

final_variable_heatmap_variables <- final_model_variable_heatmap_counts %>%
  slice_head(n = final_variable_heatmap_top_n) %>%
  mutate(
    variable_label = paste0(.data$variable, " (", .data$n_models_selected, ")")
  )

final_variable_heatmap_selected <- final_model_variable_long %>%
  distinct(model, variable) %>%
  mutate(selected = "Selected")

final_variable_heatmap_data <- tidyr::expand_grid(
  model = final_model_display_names,
  variable = final_variable_heatmap_variables$variable
) %>%
  left_join(
    final_variable_heatmap_variables,
    by = "variable"
  ) %>%
  left_join(
    final_variable_heatmap_selected,
    by = c("model", "variable")
  ) %>%
  mutate(
    selected = if_else(is.na(.data$selected), "Not selected", .data$selected),
    model = factor(.data$model, levels = final_model_display_names),
    variable_type = factor(.data$variable_type, levels = c("Demographic", "miRNA")),
    variable_label = factor(
      .data$variable_label,
      levels = rev(final_variable_heatmap_variables$variable_label)
    )
  )

final_variable_heatmap <- ggplot(
  final_variable_heatmap_data,
  aes(x = model, y = variable_label, fill = selected)
) +
  geom_tile(color = "white", linewidth = 0.6) +
  facet_grid(variable_type ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = c(
      "Not selected" = "#ece8df",
      "Selected" = "#0f5c78"
    )
  ) +
  labs(
    title = "Final Variable Selection Heatmap",
    subtitle = final_variable_heatmap_subtitle,
    x = "Model",
    y = "Variable (number of models selecting it)",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 35, hjust = 1),
    strip.text.y = element_text(angle = 0),
    legend.position = "top"
  )

final_variable_heatmap_file <- file.path("Figures", "final_variable_selection_heatmap.png")

save_plot(
  filename = final_variable_heatmap_file,
  plot = final_variable_heatmap,
  width = 8,
  height = max(5, 0.3 * nrow(final_variable_heatmap_variables) + 1.5)
)

cat(
  sprintf(
    '<img src="%s" alt="Final Variable Selection Heatmap" style="width:100%%; max-width:900px;" />\n\n',
    final_variable_heatmap_file
  )
)

```

## Calibration Curves and Additional Metrics
Reliability diagrams (calibration curves) and Brier scores for each model on the holdout set

```{r calibration_curves, echo=FALSE, results='asis', fig.width=7, fig.height=5}

calibration_summary <- holdout_predictions %>%
  group_by(.data$model) %>%
  summarise(
    brier_score = mean((.data$prob - .data$observed)^2, na.rm = TRUE),
    .groups = "drop"
  )

calibration_plot_data <- holdout_predictions %>%
  mutate(
    prob_bin = cut(
      .data$prob,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE,
      right = TRUE
    )
  ) %>%
  group_by(.data$model, .data$prob_bin) %>%
  summarise(
    mean_prob = mean(.data$prob, na.rm = TRUE),
    observed_rate = mean(.data$observed, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(.data$mean_prob), !is.na(.data$observed_rate))

calibration_plot_files <- list()

for (model_name in model_names) {
  calibration_data <- calibration_plot_data %>%
    filter(.data$model == model_name)
  brier_score <- calibration_summary %>%
    filter(.data$model == model_name) %>%
    pull("brier_score")

  calibration_plot <- ggplot(calibration_data, aes(x = mean_prob, y = observed_rate)) +
    geom_point(aes(size = n)) +
    geom_line() +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(
      title = paste("Calibration Curve -", model_name),
      subtitle = paste("Brier score =", round(brier_score, 4)),
      x = "Mean Predicted Probability",
      y = "Observed Event Rate",
      size = "Bin N"
    ) +
    theme_minimal()

  calibration_plot_file <- file.path("Figures", paste0("calibration_curve_", model_name, ".png"))
  calibration_plot_files[[model_name]] <- calibration_plot_file

  save_plot(
    filename = calibration_plot_file,
    plot = calibration_plot,
  )
}

calibration_summary <- calibration_summary %>%
  arrange(.data$brier_score)

write.csv(calibration_summary, "Tables/Calibration Summary.csv", row.names = FALSE)

render_html_table(
  calibration_summary,
  caption = "Brier scores by model",
  digits = 4
)

for (model_name in model_names) {
  cat("### ", model_display_name(model_name), "\n\n", sep = "")
  cat(
    sprintf(
      '<img src="%s" alt="Calibration Curve - %s" style="width:100%%; max-width:700px;" />\n\n',
      calibration_plot_files[[model_name]],
      model_name
    )
  )
}

```


# Interpretation and Visualization
ROC curve comparison of all models on the test set.

```{r roc_curve_comparison, echo=FALSE, results='asis', fig.width=7, fig.height=5}

# ROC curve plot with all six models on the holdout set
render_html_table(
  holdout_summary %>% arrange(desc(.data$test_auc)),
  caption = "Holdout ROC performance summary",
  digits = 3
)

roc_curves <- lapply(model_names, function(model_name) {
  roc(final_fold_inputs$valid_model_dat$CC_bin, final_model_outputs[[model_name]]$prob, quiet = TRUE)
})
names(roc_curves) <- model_names

roc_labels <- vapply(model_names, function(model_name) {
  sprintf("%s (AUC = %.3f)", model_name, as.numeric(auc(roc_curves[[model_name]])))
}, character(1))

roc_label_order <- roc_labels[holdout_summary$model[order(-holdout_summary$test_auc)]]

roc_curve_data <- bind_rows(lapply(model_names, function(model_name) {
  roc_obj <- roc_curves[[model_name]]

  data.frame(
    false_positive_rate = 1 - roc_obj$specificities,
    true_positive_rate = roc_obj$sensitivities,
    legend_label = roc_labels[[model_name]],
    row.names = NULL
  )
}))

roc_curve_data$legend_label <- factor(roc_curve_data$legend_label, levels = roc_label_order)

roc_curve_plot <- ggplot(
  roc_curve_data,
  aes(x = false_positive_rate, y = true_positive_rate, color = legend_label)
) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title = "ROC Curve Comparison of All Models",
    x = "FP Rate",
    y = "TP Rate",
    color = "Model"
  ) +
  theme_minimal()

roc_curve_plot_file <- file.path("Figures", "roc_curve_comparison.png")

save_plot(
  filename = roc_curve_plot_file,
  plot = roc_curve_plot,
)

cat(
  sprintf(
    '<img src="%s" alt="ROC Curve Comparison of All Models" style="width:100%%; max-width:700px;" />\n\n',
    roc_curve_plot_file
  )
)


```

# Selected miRNA Features and Interpretation

## Final selected miRNA across all models

```{r final_selected_miRNA, echo=FALSE, results='asis'}

# final feature counts across all models:
# create a table of all miRNA features selected by any final model, with a count of how many models selected each feature, sorted by count and then alphabetically
final_selected_miRNA <- bind_rows(lapply(model_names, function(model_name) {
  model_selected <- final_model_outputs[[model_name]]$selected_miRNA

  if (is.null(model_selected) || !length(model_selected)) {
    return(data.frame(
      model = character(0),
      miRNA = character(0),
      stringsAsFactors = FALSE
    ))
  }

  data.frame(
    model = rep(model_name, length(model_selected)),
    miRNA = model_selected,
    stringsAsFactors = FALSE
  )
}))

final_model_feature_counts <- final_selected_miRNA %>%
  distinct(model, miRNA) %>%
  count(miRNA, name = "n_models_selected") %>%
  arrange(desc(n_models_selected), miRNA)

write.csv(final_model_feature_counts, "Tables/Final Model Feature Counts.csv", row.names = FALSE)

render_html_table(
  final_model_feature_counts,
  caption = "Counts of models selecting each miRNA"
)

if (nrow(final_model_feature_counts) == 0) {
  stop("No miRNAs were selected by any final model.", call. = TRUE)
}

final_filter_stats <- final_filter_results$stats

gpl <- read_tsv(
  gene_family_file,
  col_names = c("Annotation", "Alias", "MIMAT", "ClassName"),
  show_col_types = FALSE
)

annotation_lookup <- gpl %>%
  distinct(.data$MIMAT, .keep_all = TRUE) %>%
  transmute(
    miRNA = .data$MIMAT,
    Annotation = .data$Annotation
  )

top_ranked_miRNA <- final_model_feature_counts %>%
  slice_max(order_by = n_models_selected, n = 1, with_ties = TRUE) %>%
  arrange(miRNA)
top_rank_count <- top_ranked_miRNA$n_models_selected[[1]]

top_ranked_miRNA_details <- top_ranked_miRNA %>%
  left_join(final_filter_stats[, filter_stat_cols], by = "miRNA") %>%
  left_join(annotation_lookup, by = "miRNA") %>%
  arrange(.data$limma_p_value, desc(abs(.data$limma_t)), .data$miRNA)

top_ranked_miRNA_details$multiMiR_query_id <- vapply(
  top_ranked_miRNA_details$Annotation,
  FUN.VALUE = character(1),
  function(annotation_value) {
    if (is.na(annotation_value) || !nzchar(annotation_value)) {
      return(NA_character_)
    }

    trimws(strsplit(annotation_value, ",", fixed = TRUE)[[1]][[1]])
  }
)

cat(
  "<p>Top-rank selection count: ",
  top_rank_count,
  " of ",
  length(model_names),
  " models</p>\n",
  sep = ""
)

render_html_table(
  top_ranked_miRNA_details %>%
    dplyr::select(
      miRNA,
      Annotation,
      n_models_selected,
      limma_log_fc,
      limma_t,
      limma_p_value,
      limma_adj_p_value
    ),
  caption = "Top-ranked consensus miRNAs"
)

cat(
  "<p>Downstream multiMiR analysis uses all ",
  nrow(top_ranked_miRNA_details),
  " miRNAs in this top-ranked consensus set.</p>\n",
  sep = ""
)

if (nrow(top_ranked_miRNA_details) > 1) {
  cat(
    paste0(
      "<p>Multiple miRNAs tied for the highest model-selection count, ",
      "so the table below is ordered by the smallest blocked-limma p-value, ",
      "with ties broken by the absolute moderated t-statistic and then miRNA identifier. ",
      "All rows in this table are used for downstream multiMiR annotation.</p>\n"
    )
  )
}

```

## Final selected variables across all models
```{r final_selected_variables, echo=FALSE, results='asis'}

# final feature counts across all models:
# create a table of all features selected by any final model, with a count of how many models selected each feature, sorted by count and then alphabetically
final_selected_variables <- bind_rows(lapply(model_names, function(model_name) {
  model_selected <- final_model_outputs[[model_name]]$selected_variables

  if (is.null(model_selected) || !length(model_selected)) {
    return(data.frame(
      model = character(0),
      variable = character(0),
      stringsAsFactors = FALSE
    ))
  }

  data.frame(
    model = rep(model_name, length(model_selected)),
    variable = model_selected,
    stringsAsFactors = FALSE
  )
}))

final_model_variable_counts <- final_selected_variables %>%
  distinct(model, variable) %>%
  mutate(
    variable_type = if_else(is_model_covariate_term(.data$variable), "Demographic", "miRNA")
  ) %>%
  count(variable, variable_type, name = "n_models_selected") %>%
  arrange(desc(n_models_selected), variable_type, variable)

write.csv(final_model_variable_counts, "Tables/Final Variable Counts.csv", row.names = FALSE)

render_html_table(
  final_model_variable_counts,
  caption = "Counts of models selecting each variable"
)

if (nrow(final_model_variable_counts) == 0) {
  stop("No variables were selected by any final model.", call. = TRUE)
}

final_filter_stats <- final_filter_results$stats

gpl <- read_tsv(
  gene_family_file,
  col_names = c("Annotation", "Alias", "MIMAT", "ClassName"),
  show_col_types = FALSE
)

annotation_lookup <- gpl %>%
  distinct(.data$MIMAT, .keep_all = TRUE) %>%
  transmute(
    miRNA = .data$MIMAT,
    Annotation = .data$Annotation
  )

top_ranked_variable <- final_model_variable_counts %>%
  slice_max(order_by = n_models_selected, n = 1, with_ties = TRUE) %>%
  arrange(variable_type, variable)
top_rank_count <- top_ranked_variable$n_models_selected[[1]]

top_ranked_variable_details <- top_ranked_variable %>%
  mutate(
    miRNA = if_else(.data$variable_type == "miRNA", .data$variable, NA_character_)
  ) %>%
  left_join(final_filter_stats[, filter_stat_cols], by = "miRNA") %>%
  left_join(annotation_lookup, by = "miRNA") %>%
  mutate(
    Annotation = if_else(
      .data$variable_type == "Demographic",
      "Demographic covariate",
      .data$Annotation
    )
  ) %>%
  arrange(is.na(.data$limma_p_value), .data$limma_p_value, desc(abs(.data$limma_t)), .data$variable)

top_ranked_variable_details$multiMiR_query_id <- vapply(
  top_ranked_variable_details$Annotation,
  FUN.VALUE = character(1),
  function(annotation_value) {
    if (is.na(annotation_value) || !nzchar(annotation_value)) {
      return(NA_character_)
    }

    trimws(strsplit(annotation_value, ",", fixed = TRUE)[[1]][[1]])
  }
)

top_ranked_consensus_variable <- top_ranked_variable_details %>%
  slice_head(n = 1)

cat(
  "<p>Top-rank selection count: ",
  top_rank_count,
  " of ",
  length(model_names),
  " models</p>\n",
  sep = ""
)

render_html_table(
  top_ranked_variable_details %>%
    dplyr::select(
      variable,
      variable_type,
      Annotation,
      n_models_selected,
      limma_log_fc,
      limma_t,
      limma_p_value,
      limma_adj_p_value
    ),
  caption = "Top-ranked consensus variables"
)

if (nrow(top_ranked_variable_details) > 1) {
  cat(
    paste0(
      "<p>Multiple variables tied for the highest model-selection count, ",
      "so miRNA variables in the table below are ordered by the smallest blocked-limma p-value, ",
      "with ties broken by the absolute moderated t-statistic and then variable identifier. ",
      "Demographic covariates do not have blocked-limma statistics and are listed after miRNA variables. ",
      "The first row is the top-ranked consensus variable used for downstream annotation.</p>\n"
    )
  )
}

```

## miRNA Interactions
Assess validated target and disease/drug associations for the top-ranked consensus miRNA set with multiMiR.
```{r multi_mir_analysis, eval=run_multimir_analysis, echo=FALSE, results='asis'}

multimir_results <- run_multi_mir_analysis(top_ranked_miRNA_details)

if (nrow(multimir_results$queried_miRNA) == 0) {
  cat("<p>No valid miRNA annotation was available for the top-ranked consensus miRNA set, so multiMiR queries were skipped.</p>\n")
} else {
  cat(
    "<p>multiMiR query miRNAs: ",
    write_comma_separated_list(multimir_results$queried_miRNA$multiMiR_query_id),
    "</p>\n",
    sep = ""
  )

  if (multimir_results$skipped_query_count > 0) {
    cat(
      "<p>",
      multimir_results$skipped_query_count,
      " top-ranked consensus miRNA",
      if (multimir_results$skipped_query_count == 1) " was" else "s were",
      " skipped because no valid multiMiR query identifier was available.</p>\n",
      sep = ""
    )
  }

  validated_target_summary <- multimir_results$validated$summary

  if (!is.null(validated_target_summary) && nrow(validated_target_summary) > 0) {
    write.csv(
      validated_target_summary,
      "Tables/Top-ranked Consensus miRNA multiMiR Validated Targets.csv",
      row.names = FALSE
    )

    unique_validated_targets <- unique(
      validated_target_summary$target_symbol[
        !is.na(validated_target_summary$target_symbol) &
          validated_target_summary$target_symbol != ""
      ]
    )

    cat(
      "<p>Validated multiMiR interaction records across the top-ranked consensus miRNA set: ",
      multimir_results$validated$total_record_count,
      "; unique target genes: ",
      length(unique_validated_targets),
      "; queried miRNAs with validated targets: ",
      length(unique(validated_target_summary$miRNA)),
      " of ",
      nrow(multimir_results$queried_miRNA),
      "</p>\n",
      sep = ""
    )

    render_html_table(
      utils::head(validated_target_summary, 15),
      caption = "Top validated multiMiR targets for the top-ranked consensus miRNA set"
    )
  } else {
    cat("<p>No validated multiMiR target interactions were returned for the top-ranked consensus miRNA set.</p>\n")
  }

  disease_association_summary <- multimir_results$disease_drug$summary

  if (!is.null(disease_association_summary) && nrow(disease_association_summary) > 0) {
    write.csv(
      disease_association_summary,
      "Tables/Top-ranked Consensus miRNA multiMiR Disease-Drug Associations.csv",
      row.names = FALSE
    )

    unique_association_terms <- unique(
      disease_association_summary$disease_drug[
        !is.na(disease_association_summary$disease_drug) &
          disease_association_summary$disease_drug != ""
      ]
    )

    cat(
      "<p>multiMiR disease/drug association records across the top-ranked consensus miRNA set: ",
      multimir_results$disease_drug$total_record_count,
      "; unique association terms: ",
      length(unique_association_terms),
      "; queried miRNAs with disease/drug associations: ",
      length(unique(disease_association_summary$miRNA)),
      " of ",
      nrow(multimir_results$queried_miRNA),
      "</p>\n",
      sep = ""
    )

    render_html_table(
      utils::head(disease_association_summary, 15),
      caption = "Top multiMiR disease/drug associations for the top-ranked consensus miRNA set"
    )
  } else {
    cat("<p>No multiMiR disease/drug associations were returned for the top-ranked consensus miRNA set.</p>\n")
  }
}

```

```{r store_multimir_results, eval=run_multimir_analysis, include=FALSE}

saveRDS(multimir_results, multimir_results_rds_file)

```

```{r load_multimir_results_from_disk, eval=FALSE, include=FALSE}

multimir_results <- readRDS(multimir_results_rds_file)

```

## Shared Validated multiMiR Target Genes
Summarize validated target genes across the saved multiMiR results object for the full top-ranked consensus miRNA set.

```{r multimir_gene_summary, echo=FALSE, results='asis'}

multimir_gene_summary <- NULL

validated_target_summary <- multimir_results$validated$summary

if (is.null(validated_target_summary) || nrow(validated_target_summary) == 0) {
  cat(
    "<p>The saved multiMiR results did not contain any validated target summary rows.</p>\n"
  )
} else {
  first_non_missing <- function(values) {
    values <- values[!is.na(values) & values != ""]

    if (length(values)) {
      values[[1]]
    } else {
      NA_character_
    }
  }

  multimir_gene_summary <- validated_target_summary %>%
    mutate(
      miRNA_label = if_else(
        !is.na(.data$Annotation) & .data$Annotation != "",
        .data$Annotation,
        .data$miRNA
      )
    ) %>%
    group_by(.data$target_symbol) %>%
    summarise(
      target_entrez = first_non_missing(.data$target_entrez),
      target_ensembl = first_non_missing(.data$target_ensembl),
      n_consensus_miRNA = n_distinct(.data$miRNA),
      consensus_miRNA = write_comma_separated_list(sort(unique(.data$miRNA_label))),
      total_validated_support = sum(.data$validated_support, na.rm = TRUE),
      total_support = sum(.data$total_support, na.rm = TRUE),
      max_models_selected = max(.data$n_models_selected, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(
      desc(.data$n_consensus_miRNA),
      desc(.data$total_validated_support),
      desc(.data$total_support),
      .data$target_symbol
    )

  write.csv(
    multimir_gene_summary,
    "Tables/Top-ranked Consensus miRNA multiMiR Target Gene Summary.csv",
    row.names = FALSE
  )

  cat(
    "<p>Loaded saved multiMiR results from ",
    multimir_results_rds_file,
    " and summarized ",
    nrow(multimir_gene_summary),
    " validated target genes across ",
    nrow(multimir_results$queried_miRNA),
    " queried consensus miRNAs.</p>\n",
    sep = ""
  )

  render_html_table(
    utils::head(multimir_gene_summary, 20),
    caption = "Validated target genes shared across the top-ranked consensus miRNA set"
  )
}

```

## Shared multiMiR Disease/Drug Terms
Summarize disease/drug terms across the saved multiMiR results object for the full top-ranked consensus miRNA set.

```{r multimir_disease_drug_top_terms, echo=FALSE, results='asis'}

disease_association_summary <- multimir_results$disease_drug$summary

if (is.null(disease_association_summary) || nrow(disease_association_summary) == 0) {
  cat(
    "<p>The saved multiMiR results did not contain any disease/drug association summary rows.</p>\n"
  )
} else {
  multimir_disease_drug_summary <- disease_association_summary %>%
    mutate(
      miRNA_label = if_else(
        !is.na(.data$Annotation) & .data$Annotation != "",
        .data$Annotation,
        .data$miRNA
      ),
      disease_drug = trimws(.data$disease_drug)
    ) %>%
    filter(!is.na(.data$disease_drug), .data$disease_drug != "") %>%
    group_by(.data$disease_drug) %>%
    summarise(
      n_consensus_miRNA = n_distinct(.data$miRNA),
      consensus_miRNA = write_comma_separated_list(sort(unique(.data$miRNA_label))),
      total_records = sum(.data$n_records, na.rm = TRUE),
      total_databases = sum(.data$n_databases, na.rm = TRUE),
      total_pubmed = sum(.data$n_pubmed, na.rm = TRUE),
      max_models_selected = max(.data$n_models_selected, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(
      desc(.data$n_consensus_miRNA),
      desc(.data$total_records),
      desc(.data$total_pubmed),
      .data$disease_drug
    )

  write.csv(
    multimir_disease_drug_summary,
    "Tables/Top-ranked Consensus miRNA multiMiR Disease-Drug Term Summary.csv",
    row.names = FALSE
  )

  cat(
    "<p>Loaded saved multiMiR results from ",
    multimir_results_rds_file,
    " and summarized ",
    nrow(multimir_disease_drug_summary),
    " disease/drug terms across ",
    nrow(multimir_results$queried_miRNA),
    " queried consensus miRNAs.</p>\n",
    sep = ""
  )

  render_html_table(
    multimir_disease_drug_summary,
    caption = "Disease/drug terms shared across the top-ranked consensus miRNA set"
  )
}


```

# Final Summary Table of the Top-Ranked Consensus miRNAs
A table of the top-ranked consensus miRNAs with their blocked-limma statistics, model selection counts, and annotation names.

```{r consensus_features, echo=FALSE, results='asis'}

final_feature_table <- top_ranked_miRNA_details %>%
  dplyr::select(
    miRNA,
    Annotation,
    multiMiR_query_id,
    n_models_selected,
    tidyselect::all_of(filter_stat_cols[filter_stat_cols != "miRNA"])
  )

write.csv(final_feature_table, "Tables/Final Consensus Features.csv", row.names = FALSE)

render_html_table(
  final_feature_table,
  caption = "Top-ranked consensus miRNAs summary",
  digits = 3
)

```

## Blocked-limma Summary Plot of Top-Ranked Consensus miRNAs
Blocked-limma summary of the top-ranked consensus miRNAs.
```{r final_limma_summary_plot, echo=FALSE, results='asis', fig.width=7, fig.height=5}

limma_plot_data <- top_ranked_miRNA_details %>%
  mutate(
    plot_p_value = pmax(.data$limma_p_value, .Machine$double.xmin),
    plot_label = if_else(!is.na(.data$Annotation) & .data$Annotation != "", .data$Annotation, .data$miRNA)
  )

limma_summary_plot <- ggplot(
  limma_plot_data,
  aes(x = limma_log_fc, y = -log10(plot_p_value), color = limma_t)
) +
  geom_point(size = 3) +
  geom_text(aes(label = plot_label), nudge_y = 0.1, check_overlap = TRUE, size = 3) +
  scale_color_gradient2(low = "blue", mid = "grey90", high = "red", midpoint = 0, na.value = "grey") +
  labs(
    title = "Blocked-limma Summary of Top-Ranked Consensus miRNAs",
    x = "Limma logFC",
    y = "-Log10 limma P-Value",
    color = "Limma t"
  ) +
  theme_minimal()

association_plot_file <- "Figures/Blocked Limma Summary Top Consensus miRNAs.png"

save_plot(
  filename = association_plot_file,
  plot = limma_summary_plot,
)

cat(
  sprintf(
    '<img src="%s" alt="Blocked-limma Summary of Top-Ranked Consensus miRNAs" style="width:100%%; max-width:700px;" />\n\n',
    association_plot_file
  )
)

```

```{r final_report_conclusion, echo=FALSE, results='asis'}

if (!exists("report_start_time", inherits = TRUE)) {
  cat("<p>Elapsed runtime was not available because the start timestamp was not initialized.</p>\n")
} else {
  report_end_time <- Sys.time()
  elapsed_seconds <- as.numeric(difftime(report_end_time, report_start_time, units = "secs"))

  format_elapsed_time <- function(elapsed_seconds) {
    total_seconds <- max(0L, as.integer(round(elapsed_seconds)))
    hours <- total_seconds %/% 3600L
    minutes <- (total_seconds %% 3600L) %/% 60L
    seconds <- total_seconds %% 60L

    sprintf("%02d:%02d:%02d", hours, minutes, seconds)
  }

  cat(
    paste0(
      "<p>Report runtime: ",
      format_elapsed_time(elapsed_seconds),
      " (started ",
      format(report_start_time, "%Y-%m-%d %H:%M:%S"),
      "; finished ",
      format(report_end_time, "%Y-%m-%d %H:%M:%S"),
      ").</p>\n"
    )
  )
}

```